<!-- This is one way to do it, use an include to load qmd files from elsewhere.
Just beware that if you edit hiddenCh1.qmd any live preview won't auto update,
so it's a pain when developing. You'll need to resave THIS file to force updates

{{< include Ch1/Ch1-content.qmd >}} -->

# Chapter 1 - Headings {#sec-headings}

Each row that begins with a # is a chapter heading. So if each chapter is a single file it's best to only use a single # line at the top of each document. 

## Models for binary/binomial response {#sec-binary-response}

This chapter contains sections and subsections. This is the intro. Since the chapter title begins with a `#`, sections begin with `##`, and subsections should begin with `###`.

Typical default numbering will look like this:
```
Chapter 1
Section 1.1
Subsection 1.1.1
Subsection 1.1.2
Section 1.2
Subsection 1.2.1
Subsection 1.2.2
Subsection 1.2.3
Section 1.3
```
The `crossref` option in your global `.yml` file does offer the chance to not number any chapters and just have sections in each chapter, if you really don't want chapters. 

### First subsection

Text goes here.

### Second subsection

Text goes here. And we can reference whole sections if we wish, like @sec-binary-response. Indeed if you look in @sec-links.

Let

$$D_3(y) \equiv \diff[3]{y}{x}$$

That was just an equation inside `$$` and `$$` tags. Equations must be defined like this, they can be labelled (unlike in LaTeX). You should also avoid using LaTeX environments which begin math mode, like `align`, instead use the `-ed` version `aligned` which assumes you're already inside math mode.

$$
\int_0^{\infty} x^3 \dx
$$ {#eq-AROC}

Here's some more text

* here's bullet point $x+y=z$

* another bullet

More text.

## Section title

Here's some sample text from Advanced Predictive Models.

We will begin with models for binomial responses and we will look at exploratory plots of the data, different choices of link function and hypothesis tests about terms in the model. Finally we will examine measures of goodness of fit of the model. Let's begin with an example.

```{r}
#| echo: false
#| include: false
#| label: r-fig-setup
knitr::opts_chunk$set(comment=NA, warning = FALSE, message = FALSE)
library(ggplot2)
library(knitr)
library(dplyr)
library(ROCR)
library(sjPlot)
set_theme(base = theme_grey(), theme.font = NULL, title.color = "black",
title.size = 1.2, title.align = "left", title.vjust = NULL,
geom.outline.color = NULL, geom.outline.size = 0,
geom.boxoutline.size = 0.5, geom.boxoutline.color = "black",
geom.alpha = 1, geom.linetype = 1, geom.errorbar.size = 0.7,
geom.errorbar.linetype = 1, geom.label.color = NULL,
geom.label.size = 4, geom.label.alpha = 1, geom.label.angle = 0,
axis.title.color = "grey30", axis.title.size = 1.1,
axis.title.x.vjust = NULL, axis.title.y.vjust = NULL, axis.angle.x = 0,
axis.angle.y = 0, axis.angle = NULL, axis.textcolor.x = "grey30",
axis.textcolor.y = "grey30", axis.textcolor = NULL,
axis.linecolor.x = NULL, axis.linecolor.y = NULL, axis.linecolor = NULL,
axis.line.size = 0.5, axis.textsize.x = 1, axis.textsize.y = 1,
axis.textsize = NULL, axis.tickslen = NULL, axis.tickscol = NULL,
axis.ticksmar = NULL, axis.ticksize.x = NULL, axis.ticksize.y = NULL,
panel.backcol = NULL, panel.bordercol =NULL, panel.col = NULL,
panel.major.gridcol = NULL, panel.minor.gridcol = NULL,
panel.gridcol = NULL, panel.gridcol.x = NULL, panel.gridcol.y = NULL,
panel.major.linetype = 1, panel.minor.linetype = 1, plot.backcol = NULL,
plot.bordercol = NULL, plot.col = NULL, plot.margins = NULL,
legend.pos = "right", legend.just = NULL, legend.inside = FALSE,
legend.size = 1, legend.color = "black", legend.title.size = 1,
legend.title.color = "black", legend.title.face = "bold",
legend.backgroundcol = "white", legend.bordercol = "white",
legend.item.size = NULL, legend.item.backcol = "grey90",
legend.item.bordercol = "white")
library(gridExtra)
```

<!-- Setup python via conda and a predefined quarto environment -->
```{r}
#| label: python-setup
#| code-fold: false
#| echo: false
library(reticulate)
#These next two lines need to run ONCE on your machine
#reticulate::virtualenv_create("r-quarto")
#reticulate::py_install(c("pandas","seaborn","matplotlib","numpy"), envname = "r-quarto")
reticulate::use_virtualenv("r-quarto", required = TRUE)
```

```{python}
#| label: python-package-loading
#| echo: false
# Standard plot settings
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
import statsmodels.api as sm
import statsmodels.formula.api as smf

# This is how to control plot sizes later, define plot size before creating plot
# fig, ax = plt.subplots(figsize=(8, 5))
```

:::{.callout-tip .example}
### Beetle mortality

This is standard Quarto a "Tip Callout" box called via
```
:::{.callout-tip}
### Banner title like Bettle Mortality

Content here. This gives coloured boxes in HTML and LaTeX.
:::
```

**However**,  see @sec-niceboxes for my recommended alternative to these callout boxes.

Now for a default code `panelset` as Quarto calls it. For allowing dual code presentation. See @sec-code-demo for my restyled version.

::::::{.panel-tabset}

### R

```{r}
#| label: r-load-beetles
#| code-fold: show
beetles <- read.csv(url("http://www.stats.gla.ac.uk/~tereza/rp/beetles.csv"))
beetles
```

### Python

```{python}
#| label: python-load-beetles
#| eval: false
import pandas as pd
beetles = pd.read_csv("beetles.csv")
beetles
```
This Python code was told not to evaluate. Note it calls a local csv file, which would need to be in the right place. Paths will be relative to the `.qmd` file being compiled.

::::::

:::

