---
filters:
    - acc-tools
---

# Week 2 -- Code demonstrations {#sec-code-demo}

<!-- # Models for binary/binomial response  -->

Some equations, again. Notice this chapter begins with:
```
---
filters:
    - acc-tools
---
```

This loads the bonus accessibility features extension. There is a global code toggle button bottom left, and the R and P keyboard buttons will switch **all** code tabs simultaneously to the chosen language.

$$\diff[3]{y}{x}$$

$$\int_0^{\infty} x^2 \dx$$

This equation is the same as @eq-AROC from Chapter 1. Since we're in the book template, and all files are compiled simultaneously into one output we are allowed to link between chapters. This is the fundamental difference between the booktemplate and the weektemplate.

## Features

- You'll first note in the bottom left is a global toggle button set. You can change all code blocks to R or to Python, by clicking; or by pressing `R` or `P` on the keyboard.
- There tabs labelled R and Python have been moved to the right, and when selected (for example by pressing tab) are much more clearly selected.
- Also the copy-code button is always visible.

## New Section Title

Some more placeholder text here. Even in the book template every `.qmd` that wishes to use python needs to load it via `reticulate`. This is in the template:

<!-- Setup python via conda and a predefined quarto environment -->
```{r}
#| label: python-setup
#| echo: fenced
library(reticulate)
#These next two lines need to run ONCE on your machine
#reticulate::virtualenv_create("r-quarto")
#reticulate::py_install(c("pandas","seaborn","matplotlib","numpy"), envname = "r-quarto")
reticulate::use_virtualenv("r-quarto", required = TRUE)
```

```{python}
#| label:  py-imports
#| echo: false
#| eval: true # Note we are not evaluating, we already loaded in earlier qmd
# Standard plot settings
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
import statsmodels.api as sm
import statsmodels.formula.api as smf

# This is how to control plot sizes later, define plot size before creating plot
# fig, ax = plt.subplots(figsize=(8, 5))
```

Another standard callout box.

:::{.callout-tip}
### Beetle mortality

Jump to the [Beetle code](#beetle-code).

Data for this example consists of the number of beetles dead (*killed*) after five hours exposure to gaseous carbon disulphide at various concentrations (*dose*). The goal for this analysis is to model the probability of a beetle dying as a function of the carbon disulphide dose. 

<a id="beetle-code"></a>

::::::{.panel-tabset}

### R

The code in the R example is folded by default. Click on `Code` to reveal it.

```{r}
#| label: r-beetles-load
#| code-fold: true
beetles <- read.csv(url("http://www.stats.gla.ac.uk/~tereza/rp/beetles.csv"))
beetles
```

### Python

```{python}
#| label: py-beetles-load
import pandas as pd
beetles = pd.read_csv("resources/data/beetles.csv")
beetles
```

::::::

:::