---
filters:
    - acc-tools
---

# Week 2 -- Code demonstrations {#sec-code-demo}

<!-- # Models for binary/binomial response  -->

Some equations, again. Notice this chapter begins with:
```
---
filters:
    - acc-tools
---
```
This loads the bonus accessibility features extension. There is a global code toggle button bottom left, and the R and P keyboard buttons will switch **all** code tabs simultaneously to the chosen language.

$$\diff[3]{y}{x}$$

$$\int_0^{\infty} x^2 \dx$$

This equation is the same as @eq-AROC from Chapter 1.

## Features

- You'll first note in the bottom left is a global toggle button set. You can change all code blocks to R or to Python, by clicking; or by pressing `R` or `P` on the keyboard.

## Binomial response

We will begin with models for binomial responses and we will look at exploratory plots of the data, different choices of link function and hypothesis tests about terms in the model. Finally we will examine measures of goodness of fit of the model. Let's begin with an example.

```{r r-setup,echo=FALSE, include=FALSE, fig.height=4, fig.align='center'}
knitr::opts_chunk$set(comment=NA, warning = FALSE, message = FALSE)
library(ggplot2)
library(knitr)
library(dplyr)
library(ROCR)
library(sjPlot)
set_theme(base = theme_grey(), theme.font = NULL, title.color = "black",
title.size = 1.2, title.align = "left", title.vjust = NULL,
geom.outline.color = NULL, geom.outline.size = 0,
geom.boxoutline.size = 0.5, geom.boxoutline.color = "black",
geom.alpha = 1, geom.linetype = 1, geom.errorbar.size = 0.7,
geom.errorbar.linetype = 1, geom.label.color = NULL,
geom.label.size = 4, geom.label.alpha = 1, geom.label.angle = 0,
axis.title.color = "grey30", axis.title.size = 1.1,
axis.title.x.vjust = NULL, axis.title.y.vjust = NULL, axis.angle.x = 0,
axis.angle.y = 0, axis.angle = NULL, axis.textcolor.x = "grey30",
axis.textcolor.y = "grey30", axis.textcolor = NULL,
axis.linecolor.x = NULL, axis.linecolor.y = NULL, axis.linecolor = NULL,
axis.line.size = 0.5, axis.textsize.x = 1, axis.textsize.y = 1,
axis.textsize = NULL, axis.tickslen = NULL, axis.tickscol = NULL,
axis.ticksmar = NULL, axis.ticksize.x = NULL, axis.ticksize.y = NULL,
panel.backcol = NULL, panel.bordercol =NULL, panel.col = NULL,
panel.major.gridcol = NULL, panel.minor.gridcol = NULL,
panel.gridcol = NULL, panel.gridcol.x = NULL, panel.gridcol.y = NULL,
panel.major.linetype = 1, panel.minor.linetype = 1, plot.backcol = NULL,
plot.bordercol = NULL, plot.col = NULL, plot.margins = NULL,
legend.pos = "right", legend.just = NULL, legend.inside = FALSE,
legend.size = 1, legend.color = "black", legend.title.size = 1,
legend.title.color = "black", legend.title.face = "bold",
legend.backgroundcol = "white", legend.bordercol = "white",
legend.item.size = NULL, legend.item.backcol = "grey90",
legend.item.bordercol = "white")
library(gridExtra)
```

<!-- Setup python via conda and a predefined quarto environment -->
```{r retic-setup, echo=F}
#| code-fold: false
library(reticulate)
#reticulate::use_condaenv(condaenv = "quarto", conda = "auto", required = TRUE)
```

```{python py-imports, echo=F, eval=F}
# Standard plot settings
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
import statsmodels.api as sm
import statsmodels.formula.api as smf

# This is how to control plot sizes later, define plot size before creating plot
# fig, ax = plt.subplots(figsize=(8, 5))
```


:::{.callout-tip}
### Beetle mortality

Jump to the [Beetle code](#beetle-code).

Data for this example consists of the number of beetles dead (*killed*) after five hours exposure to gaseous carbon disulphide at various concentrations (*dose*). The goal for this analysis is to model the probability of a beetle dying as a function of the carbon disulphide dose. 

<a id="beetle-code"></a>

::::::{.panel-tabset}

### R

```{r, results = 'hide'}
#| code-fold: show
beetles <- read.csv(url("http://www.stats.gla.ac.uk/~tereza/rp/beetles.csv"))
beetles
```

### Python

```{python py-beetle-load, results = 'hide', eval=F}
import pandas as pd
# beetles = pd.read_csv("resources/data/beetles.csv")
# beetles
```

::::::

:::

:::{.callout-tip}
## Conventional tip callout

Some content
:::

:::{.callout-warning}
## Conventional warning callout

Some content
:::

:::{.callout-caution}
## Conventional caution callout

Some content
:::

:::{.callout-note}
## Conventional note callout

Some content
:::


