---
filters:
  - acc-tools
  
acc-tools:
  global-toggle: false
---

```{r python-setup-new-reticulate}
#| include: false
#| eval: true
library(reticulate)
#These next two lines need to run ONCE on your machine
#reticulate::virtualenv_create("r-quarto")
#reticulate::py_install(c("pandas","seaborn","matplotlib","numpy"), envname = "r-quarto")
reticulate::use_virtualenv("r-quarto", required = TRUE)
```

# Chapter 4 - Nicer boxes and Python {#sec-niceboxes}

This chapter shows off some nice coloured boxes. They come from the `custom-numbered-blocks` extension. First you need to install the extension `quarto add ute/custom-numbered-blocks`.

The following colour definitions need to be added somewhere (e.g. `_quarto.yml` or `_metadata.yml` or at the top of this `.qmd` file) to load the coloured boxes. In this template I have added a link to a file called `_numbered-boxes.yml` from inside `_quarto.yml`. In this way you just need to look insider `_numbered-boxes.yml` to edit the settings.

It looks approximately like this, you define groups of like-numbered boxes, and then define specific classes (and their names later). They can have separate colour schemes too.

```
filters:
  - custom-numbered-blocks # This loads the extension

# This is ute-hahn's nice coloured boxes: setup
custom-numbered-blocks:
  groups:
    thmlike:
      colors: [948bde, 584eab]
      boxstyle: foldbox.simple
    deflike:
      colors: [ffdc36, ffb948]
      boxstyle: foldbox.simple
    videolike:
      colors: [beb9eb, 877eda]
      boxstyle: foldbox.simple
      collapse: open
  classes:
    Theorem:
      group: thmlike
    Lemma:
      group: thmlike
    Corollary:
      group: thmlike
    Definition:
      group: deflike
    Video:
      group: videolike
      
crossref:
  chapters: true
```

## Custom Numbered Blocks

We’ve defined several custom-numbered block types using the `custom-numbered-blocks` option. The key point is that these can be numbered and follow grouped numbering conventions.

Each block belongs to a **group** that determines its shared numbering. 

All of these blocks are *foldable*, meaning they can be opened and closed. The `collapse` setting determines the starting state. Most boxes have been set to be open (`collapse: false`) at the group level, but this can be overridden at both lower levels. e.g. Corollary shares numbering with all the theorem-like boxes, which are all open by default. Corollaries could be all set to closed by default, or even on a case by case basis in the code.

<style>
.colorbox {
  display: inline-block;
  width: 1em;
  height: 1em;
  border: 1px solid #ccc;
  border-radius: 2px;
  margin-right: 0.3em;
  vertical-align: middle;
}
</style>

| Box Name | Group | Title Background Color | Left Border Color | Default State |
|-----------|--------|------------------|------------------------|----------------|
| Theorem   | `thmlike`  | <span class="colorbox" style="background-color:#948bde;"></span>`#948bde` | <span class="colorbox" style="background-color:#584eab;"></span>`#584eab` | Open / Closeable |
| Lemma     | `thmlike`  | <span class="colorbox" style="background-color:#b2dac4;"></span>`#b2dac4` | <span class="colorbox" style="background-color:#004721;"></span>`#004721` | Open / Closeable |
| Corollary | `thmlike`  | <span class="colorbox" style="background-color:#7fbad7;"></span>`#7fbad7` | <span class="colorbox" style="background-color:#005398;"></span>`#005398` | Open / Closeable |
| Definition| `deflike`   | <span class="colorbox" style="background-color:#ffdc36;"></span>`#ffdc36` | <span class="colorbox" style="background-color:#ffb948;"></span>`#ffb948` | Open / Closeable |
| Video     | `videolike` | <span class="colorbox" style="background-color:#d18888;"></span>`#d18888` | <span class="colorbox" style="background-color:#660000;"></span>`#660000` | Closed / Openable |

: This is the table's caption {#tbl-boxcols tbl-colwidths="[15,20,20,20,25]"}

These boxes also have dark-mode versions, see below. Notice this table had a caption, and it can be referenced with normal Quarto syntax: @tbl-boxcols. They can also have styling added.

> Note:
> The two colors defined for each group represent the **left border** and the **title background**.

Here's an alternative to treating videos as float/figures, just stick them directly onto the page immediately (not as a floating object) but wrap them in a coloured box and use the title of the box to describe the video.

:::{.Video #vidFirst}
## Binomial response models applied to beetle mortality data (9m59)  

There is some text here before the video

{{< video https://youtu.be/d2nG7Y17sQw 
  title = "Binomial response models applied to beetle mortality data (9m59s)"
>}}
:::

:::{.Example}
### Beetle mortality data

Here's some famous beetle data. The R code is folded (and thus hideable), the Python code isn't folded so cannot be folded away.

::::::{.panel-tabset}

### R

```{r r-beetle-load}
#| code-fold: show
beetles <- read.csv(url("http://www.stats.gla.ac.uk/~tereza/rp/beetles.csv"))
beetles
```

### Python

```{python py-beetle-load, eval=TRUE}
import pandas as pd
beetles = pd.read_csv("resources/data/beetles.csv")
beetles
```

::::::

:::

:::{.callout-tip}
## Conventional tip callout

Some content
:::

:::{.Corollary #bestcor}
## Super

This is a fantastic Corollary.
:::

There are no Corollary's better than Corollary \ref{bestcor}. Note that referencing of these coloured boxes works like LaTeX's `\ref{}` command, thus it just renders the number, and you have to write the object name in front. This is in contrast to the `@cor-name` Quarto notation. 

Full documentation on these coloured boxes is available at [https://github.com/ute/custom-numbered-blocks](https://github.com/ute/custom-numbered-blocks). Including the new `\longref{}` syntax allowing the object name to be included in the reference. Like this \longref{bestcor}.

The other way to embed a video is inside a fenced div and treat it like a figure, as we saw in other chapters.

Our video in this chapter was embedded inside a custom numbered block and it was called [Video \ref{vidFirst}](#vidFirst), in a moment we'll include another video. It will be called \longref{Rvid-yt}.

## Demonstration of Custom Numbered Blocks

Below are examples of the custom-numbered blocks.  
Each block is foldable (open or closed by default) and has its own colour scheme and group style as defined in your YAML configuration.

Notice Theorem, Corollary and Lemma has been defined to share a counter. Unlike Video and Definition which are different types (*vidlike* and *deflike*).

::: {.Theorem #rthmpythag}
## Pythagoras' Theorem

In a right-angled triangle...
:::

::: {.Corollary #Rcor-pythag}
## Corollary to Pythagoras' Theorem

In a right-angled triangle, if the two shorted sides are equal, so that $a = b$, then the hypotenuse satisfies $c=a\sqrt{2}$.
:::

::: {.Lemma #Rlem-euclid}
## Euclid

If two lines are perpendicular to the same line, then they are parallel to each other.
:::

::: {.Video #Rvid-yt}
## Video Demonstration
 
Here's a Youtube video:

{{< video https://youtu.be/d2nG7Y17sQw 
  title = "Binomial response models applied to beetle mortality data (9m59s)"
>}}
:::

::: {.Definition #Rdef-orthog} 
## Definition of Orthogonality

Two vectors $\mathbf{u}$ and $\mathbf{v}$ are said to be *orthogonal* if their dot product is zero:
$$
\mathbf{u} \cdot \mathbf{v} = 0.
$$
:::

---

## Referencing the Blocks

You can reference the blocks defined above using their labels:

- The main result is stated in Theorem \ref{rthmpythag}. 
- Checkout Corollary \ref{Rcor-pythag}.  
- The proof uses Lemma \ref{Rlem-euclid}.  
- Then there's Definition \ref{Rdef-orthog}.  
- Finally, see Video \ref{Rvid-yt}.

If we want we can wrap the `\ref{}` inside a hyperlink to get [Theorem \ref{rthmpythag}](#rthmpythag), [Corollary \ref{Rcor-pythag}](#Rcor-pythag), etc...

*Update*: The `custom-numbered-blocks` extension now offers `\longref{}` as a way to make the hyperlink include the word of the type of object, like `\longref{rthmpythag}` now rendering as \longref{rthmpythag}.

## Customization

Each box's definition uses two colours. In the file mentioned above the light-mode border and background are set. For those using dark-mode you also need to define a dark mode pair (so you'll need four total colours). See [dark styles boxes stylesheet](../themes/dark-styles-boxes.scss) for how to define them as it's not available as a default setting.

## More Python

Rendering documents containing R and Python is based on the `reticulate` and `knitr` R packages.

One way is to let Quarto manage a Python environment for you via `quarto install python`, but if you already have Python installed, the recommended approach now is to use a **persistent virtual environment**.

### Setting up a persistent Python environment

Look at the top of each file and you'll find a chunk like this

```{{r python-setup-new-reticulate}}
#| include: false
#| eval: true
library(reticulate)

# These next two lines need to run ONCE on your machine to create the environment
# reticulate::virtualenv_create("r-quarto")
# reticulate::py_install(c("pandas","seaborn","matplotlib","numpy"), envname = "r-quarto")

# Use the virtual environment in this document
reticulate::use_virtualenv("r-quarto", required = TRUE)
```

* **First time only:** Uncomment the `virtualenv_create` and `py_install` lines to set up the environment.
* **After setup:** Only the `use_virtualenv` line is needed.

The advantage of this approach is that libraries are installed once and persist across sessions, so you don’t incur installation overhead every time you render the document.

If while developing your notes you realize you need to load another library, then re-run the `py_install` line to install this library into your `r-quarto` environment, e.g. by running these lines in the console:

```markdown
library(reticulate)
reticulate::py_install(c("insert-new-library-here"), envname = "r-quarto")
```

### Using Python in your Quarto document

Once the environment is set, you can run Python chunks as usual:

This will demonstrate the R/Python switcher again. Notice this page doesn't contain the global toggle button in the bottom left as it was disabled at the top of the `.qmd`, should you wish to disable it.

:::{.panel-tabset} 

### R
```{r}
#| label: fig-rnorm
#| fig-cap: "A base R plot"
#| fig-alt: "Alt text"

x <- rnorm(50)
plot(x)
```

### Python

```{python eval=TRUE}
#| label: fig-polar
#| fig-cap: "A line plot on a polar axis"
#| fig-alt: "Alt text"

import numpy as np
import matplotlib.pyplot as plt

r = np.arange(0, 2, 0.01)
theta = 2 * np.pi * r
fig, ax = plt.subplots(
  subplot_kw = {'projection': 'polar'} 
)
ax.plot(theta, r)
ax.set_rticks([0.5, 1, 1.5, 2])
ax.grid(True)
plt.show()
```

:::

:::{.panel-tabset}

### R

```{r}
#| label: fig-r-cubic-plot
#| fig-cap: $f(x)=x^3$ between $-10$ and $+10$
#| fig-alt: "A plot of x-cubed between -10 and 10"
#| 
x <- seq(-10,10, by = 0.1)
y <- x ^ 3
plot(x, y, type = "l")
```

### Python
```{python}
#| label: fig-py-cubic-plot
#| fig-cap: $f(x)=x^3$ between $-10$ and $+10$
#| fig-alt: "A plot of x-cubed between -10 and 10"
#| 
import matplotlib.pyplot as plt
import numpy as np

xpoints = np.arange(-10, 10, 0.1)
ypoints = np.power(xpoints, 3)

plt.plot(xpoints, ypoints)
plt.show()
```

:::

The output images from these pieces of code have their own labels namely `@fig-rnorm` and `@fig-polar` which reference as @fig-rnorm and @fig-polar.

