---
filters:
    - acc-tools
---

# Introduction to Generalised Linear Models

:::{.Outcomes .unnumbered}

* See how a generalised linear model differs from a linear model, including the approach taken in logistic regression

* Meet the exponential family of distributions

* Be able to identify the key components of a generalised linear model (GLM)

* Appreciate the use of likelihoods, maximum likelihood estimation, and the definition of deviance in GLM modelling

* Appreciate the consequences of using maximum likelihood estimation for inference, including confidence intervals, and the use of deviance in likelihood ratio tests

*These outcomes are further reinforced with concrete examples in coming weeks.*

:::

## Introduction and motivating examples

```{r}
#| label: r_global_chunk_settings_and_lib_load
#| include: false

knitr::opts_chunk$set(comment=NA, warning = FALSE)
library(ggplot2)
```

```{r}
#| label: python-setup-real
#| echo: false
#| 
library(reticulate)

#These next two lines need to run ONCE on your machine
#reticulate::virtualenv_create("r-quarto")
#reticulate::py_install(c("pandas","seaborn","matplotlib","numpy","statsmodels"), envname = "r-quarto")
reticulate::use_virtualenv("r-quarto", required = TRUE)
```

<!-- Hide Bollywood ODL video 
:::{.Video #rvid-vid5u1w6eROypI}
## Introduction to GLMs (9m57s)
{{< video https://youtu.be/5u1w6eROypI
title = "Introduction to GLMs (9m57s)"
>}}
::: -->
<!-- end of vid -->

<!-- In this course we will extend the theory of linear regression models, covered in [Predictive Modelling](http://moodle2.gla.ac.uk/course/view.php?id=12738) and [Learning from Data](http://moodle2.gla.ac.uk/course/view.php?id=12736).  -->

In this course we will extend the theory of linear regression models to non-normal responses. Before we begin with introducing the class of models known as Generalised Linear Models (GLMs), we will briefly illustrate why linear models are not sufficient for all types of data. Throughout the course, we will see how we can deal with a variety of situations where the linear model may not be adequate.

The main objective of this week's learning material is to introduce Generalised Linear Models (GLMs), which extend the linear model framework to outcome variables that don't follow the normal distribution. GLMs can be used to model non-normal continuous outcome variables, but they are most frequently used to model [binary]{.term}, [categorical]{.term} or [count data]{.term}. We will focus on these latter types of outcome variables. To see why extensions to the normal linear model are needed, let's look at a couple of examples, one where the normal linear model is appropriate and one where it's not.



:::{.Example}

## Bollywood box office revenue

Possibly the simplest scenario of a predictive model is when we want to predict an outcome variable based on a predictor which displays a linear relationship to the variable of interest. Consider the following dataset on Bollywood film revenues (sourced from [http://www.bollymoviereviewz.com](http://www.bollymoviereviewz.com)) which contains data on 190 films made during the period 2013-2017. We would like to predict the gross revenue of a film from the film's budget. Both the gross revenue and the budget are measured in [crore](https://en.wikipedia.org/wiki/Crore). Here are the first few rows of the data:

\pagebreak

:::{.panel-tabset}

### R

```{r}
#| label: r_load_bollywood
bollywood <- read.csv('https://github.com/UofGAnalyticsData/APM/raw/refs/heads/main/bollywood_boxoffice.csv', fileEncoding = "latin1")
head(bollywood)
```

### Python

```{python}
#| label: python_load_bollywood
#| eval: true
import pandas as pd
bollywood = pd.read_csv("resources/ForChX/bollywood_boxoffice.csv", encoding = 'latin-1')
bollywood.head()
``` 

<!--    Movie                            Gross       Budget -->
<!--    -------------------------------- ----------- ----------- -->
<!--      Ek Villain                       95.64       36.0 -->
<!--      Humshakals                       55.65       77.0 -->
<!--      Holiday                         110.01      90.0 -->
<!--      Fugly                            11.16       16.0 -->
<!--      City Lights                       5.19        9.5 -->
<!--      Kuku Mathur Ki Jhand Ho Gayi      2.23        4.5 -->

:::

We can plot the gross revenue against the budget to explore the relationship between the two variables.

:::{.panel-tabset}

```{python}
#| label: python-prepare-budget-gross
#| include: false
#| eval: true
import pandas as pd
import seaborn as sns
import matplotlib.pyplot as plt
sns.set(style="white")
plt.figure(figsize=(8, 6))
```

### R

```{r}
#| label: r-prepare-budget-gross
#| results: 'hide'
b.plot <- ggplot(data = bollywood, aes(y = Gross, x = Budget)) +
          geom_point(col = "#66a61e") +
          scale_x_continuous("Budget (crore)") + scale_y_continuous("Gross (crore)")
```

```{r}
#| echo: false
#| label: r-budget-gross
#| fig-alt: "Scatterplot of budget against gross for films. Showing a weak positive correlation. No values top left, more outliers at large budgets."
b.plot + theme(panel.background = element_rect(fill = "transparent", colour = NA),
               plot.background = element_rect(fill = "transparent", colour = NA),
               panel.border = element_rect(fill = NA, colour = "black", size = 1))
```

### Python

```{python}
#| label: python-budget-gross
#| fig-cap: 'Hello'
#| fig-alt: 'Scatterplot of budget against gross for films. Showing a weak positive correlation. No values top left, more outliers at large budgets.'
#| 
sns.scatterplot(data=bollywood, x='Budget', y='Gross', color="#66a61e")
plt.xlabel("Budget (crore)")
plt.ylabel("Gross (crore)")
plt.show()
```
:::

Looking at the scale of the values on both the horizontal and vertical axes, we might want to transform the data by taking logs.

:::{.panel-tabset}

### R

```{r, results='hide', fig.height=4}
#| label: r-budget-gross-log10
b.plot.l <- ggplot(data = bollywood, aes(y = log10(Gross), x = log10(Budget))) +
          geom_point(col = "#1b9e77") +
          scale_x_continuous("log(Budget) (crore)") +
          scale_y_continuous("log(Gross) (crore)")
```

```{r, echo=FALSE, fig.height=4.5}
#| label: r-budget-gross-log10-theme
#| fig-alt: "Scatterplot of log of budget against log of gross, for films. Clear positive correlation with wide band of points stretching from lower left to top right. Few to no values top left or bottom right."
#| 
b.plot.l + theme(panel.background = element_rect(fill = "transparent", colour = NA),
                 plot.background = element_rect(fill = "transparent", colour = NA),
                 panel.border = element_rect(fill = NA, colour = "black", linewidth = 1))
```

### Python

```{python}
#| label: python-budget-gross-log10
#| fig-alt: "Scatterplot of log of budget against log of gross, for films. Clear positive correlation with wide band of points stretching from lower left to top right. Few to no values top left or bottom right."
#| eval: true

import numpy as np

bollywood['log_Budget'] = np.log10(bollywood['Budget'])
bollywood['log_Gross'] = np.log10(bollywood['Gross'])

sns.set_style("white")
sns.scatterplot(data=bollywood, x='log_Budget', y='log_Gross', color="#66a61e")
plt.xlabel("log10(Budget) (crore)")
plt.ylabel("log10(Gross) (crore)")
plt.title("Log-Log Plot of Budget vs Gross")
plt.tight_layout()
plt.show()
```

:::

Now let's fit a model with the $\log_{10}$ transformed gross revenue as the response ($Y_{i}$)  and the $\log_{10}$ transformed budget ($x_{i}$) as the explanatory/predictor variable. 

:::{.panel-tabset}

### R

We can use the `lm()` function to fit this linear model in R.

```{r}
#| label: r-lm-log10-gross-budget
bol.lm <- lm(log10(Gross) ~ log10(Budget), data = bollywood)
```

### Python

```{python}
#| label: python-lm-log10-gross-budget
import statsmodels.formula.api as smf
bol_lm = smf.ols('np.log10(Gross) ~ np.log10(Budget)', data=bollywood).fit()
```

:::

The model equation in mathematical notation is
$$E(Y_{i}) = \mu_i= \beta_0 + \beta_1 x_{i}; ~~~ \text{where the}~ Y_{i} \text{ are independent }  N(\mu_{i}, \sigma^2),~~~i=1,...,190 $$
The model fit is shown below:

:::{.panel-tabset}

### R

```{r}
#| label: r-lm-log10-gross-budget-summary
summary(bol.lm)
```

### Python

```{python}
#| label: python-lm-log10-gross-budget-summary
print(bol_lm.summary2())
```
:::

We can visualise this regression model by plotting the data and fitted regression line:

:::{.panel-tabset}

### R

```{r, results='hide', fig.height=4.5}
#| label: r-lm-log10-gross-budget-bestfit
b.plot.lm <- ggplot(data = bollywood, aes(y = log10(Gross), x = log10(Budget))) +
          geom_point(col ="#1b9e77") +
          scale_x_continuous("log(Budget) (crore)") +
          scale_y_continuous("log(Gross) (crore)") +
          geom_smooth(method = lm, colour="#e7298a", se=FALSE)
```

```{r, fig.height=4.5}
#| label: r-lm-log10-gross-budget-bestfit-theme
#| fig-alt: "Same catterplot of log of budget against log of gross, from earlier. Now with the best fit line drawn, well fitting through centre of wide band of points from top left to bottom right."
#| echo: false
b.plot.lm + theme(panel.background = element_rect(fill = "transparent", colour = NA),
                 plot.background = element_rect(fill = "transparent", colour = NA),
                 panel.border = element_rect(fill = NA, colour = "black", size = 1))
```

### Python

```{python}
#| label: python-lm-log10-gross-budget-bestfit
#| fig-alt: "Same catterplot of log of budget against log of gross, from earlier. Now with the best fit line drawn, well fitting through centre of wide band of points from top left to bottom right."

sns.set_theme(style="white")
g = sns.lmplot(
    data=bollywood,
    x='log_Budget',
    y='log_Gross',
    scatter_kws={'color': '#1b9e77'},
    line_kws={'color': '#e7298a'},
    ci=None,
    aspect=2
)

g.set_axis_labels("log(Budget) (crore)", "log(Gross) (crore)");

plt.show()
```
:::

:::  
<!-- end of Example -->


:::{.Task}
Using the fitted model equation, predict the gross revenue for a film with a budget of (i) 10, (ii) 50, and (iii) 100 crore.

*Hint: Remember that the variables have been log-transformed.*
:::
<!-- end of Task -->

::::{.Answer}
For the *Bollywood box office revenue* example, we can write down the fitted model equation from the `summary(bol.lm)`:
$$\log_{10}(\text{Gross}) =-0.62549 + 1.31955\times \log_{10}(\text{Budget})$$

We can use this equation to predict the gross revenue of a film by simply substituting the relevant budget value, and transforming the result from the log10 scale. Thus:

(i) budget = 10:
 $$
 \log_{10}(\text{Gross}) =-0.62549 + 1.31955\times \log_{10}(10) =  0.69406
 $$
 $$
 \Rightarrow \text{Gross}= 10^{0.69406} = 4.94379
 $$

(ii) budget = 50:
 $$
 \log_{10}(\text{Gross}) =-0.62549 + 1.31955\times \log_{10}(50) =  1.616386
 $$
 $$
 \text{Gross}= 10^{1.616386} = 41.34148
 $$


(iii) budget = 100:
$$
\log_{10}(\text{Gross}) =-0.62549 + 1.31955\times \log_{10}(100) =  2.01361
$$
$$
\Rightarrow \text{Gross}= 10^{2.01361} = 103.1834
$$
  
::::
<!-- end of Answer -->

:::{.Example}

## GPA and admission to medical school

Now let's look at a different kind of dataset, where the outcome we want to predict is not continuous-valued but binary. This is a dataset on admissions to US medical schools, which gives the  admission status, GPA and standardised test scores for 55 medical school applicants from a liberal arts college in the US Midwest and it can be loaded from the `Stat2Data` package in R.



<!-- This is a dataset on admissions to US medical schools which you have first seen in [Predictive Modelling](http://moodle2.gla.ac.uk/pluginfile.php/1457608/mod_resource/content/0/week9.pdf). -->


```{r}
#| label: r-load-MedGPA
library(Stat2Data)
data(MedGPA)
```

```{python}
#| include: false # Don't show this
#| label: python-load-MedGPA

MedGPA = pd.read_csv("resources/ForChX/MedGPA.csv")
MedGPA.head()
```

The first few rows of the data are given below.
```{r, echo=FALSE}
# knitr::kable(head(MedGPA), format = "markdown", padding = 2)
#| label: r-load-medGPA-packages
head(MedGPA)
library(ggplot2)
library(ggthemes)
```

Let us look at a plot of acceptance against GPA, adding a bit of jitter to make overlapping points more visible.

```{r, fig.height=4.5}
#| label: r-medgpa-scatterplot
medgpa.plot <- ggplot(data = MedGPA, aes(y = Acceptance, x = GPA)) +
               geom_jitter(width =0, height =0.01, alpha =0.5, colour ="#984ea3")
```

```{r, echo=FALSE, fig.height=4.5}
#| label: r-medgpa-scatterplot-theming
medgpa.plot <- medgpa.plot +
               theme(panel.background = element_rect(fill = "transparent", colour = NA),
               plot.background = element_rect(fill = "transparent", colour = NA),
               panel.border = element_rect(fill = NA, colour = "black", size = 1))
```

We can add the linear regression line for `Acceptance` as a function of `GPA` to the plot.


```{r, fig.height=4.5}
#| label: r-medgpa-scatterplot-lm-line
#| fig-alt: "Scatterplot of GPA against acceptance, all points at height 0 or 1, scattered from GPA 2.5 to 4.0. Best fit line drawn bottom left to top right, no fit as only two discrete y-axis values. More points on left at 0, more on right at 1."
medgpa.plot + geom_smooth(method = "lm", se = FALSE,
                          fullrange = TRUE, colour = "#984ea3")
```


The R code for fitting the model and the model output is shown below.

```{r}
#| label: r-medgpa-lm-summary
med.lm <- lm(Acceptance ~ GPA, data=MedGPA)
summary(med.lm)
```

In mathematical notation, the value of the independent response $Y_i$ is equal to 1 if the $i$th applicant is accepted and $Y_i=0$ otherwise, where $x_i$ refers to to the $i$th applicant's college GPA for $i=1,\dots,55$. The normal linear model assumes that the $Y_i$ are independent $N(\mu_i,\sigma^2)$ with $\mu_i=\beta_0+\beta_1 x_i$, with the fitted model equation given by $\hat{\mu}_i=-2.8240+0.9483x_i$.


One issue with this fit is that the predicted values of the response can take any real values, while acceptance can only take the value 0 or 1. And it is hard to argue that a variable taking only values 0 and 1 can be reasonably assumed to be normally distributed.

:::
<!-- end of example -->

:::{.Key .unnumbered}

## Mini introduction to generalised linear models

When faced with response variables which are clearly not normally distributed, and not even approximately normal, we take a multi-level approach. 

We will introduce a [logistic regression model]{.term} to target the [probability of acceptance]{.term}. To do this we will setup our modelling a specfic way, making certain assumptions.

We begin by writing $p_i=P(Y_i=1)$ to denote the probability of acceptance for the $i$th applicant. Then we assume that the $Y_i$ are independent random variables which follow the $\text{Bin}(1,p_i)$ (or Bernoulli($p_i)$) distribution. Our big idea is that our regression model will choose $p_i$ to be determined by the following equation:

$$
\log \left(\dfrac{p_i}{1-p_i} \right)=\beta_0+\beta_1 x_i,
$$

i.e. we are still using a linear model underneath, but no longer directly fitting to the raw response variable values. The choice of function on the left will later be called a [link function]{.term}.

After some algebraic re-arrangement, this is equivalent to:

$$
p_i=\dfrac{\exp(\beta_0+\beta_1 x_i)}{1+\exp(\beta_0+\beta_1 x_i)}.
$$

(*You should check you know how to do this!*)

:::

Now we can return and apply this idea to our GPA data example.

:::{.Example}

:::{.panel-tabset}

### R

We can fit this model in R using the `glm()` function:

```{r}
#| label: r-fit-glm-gpa
med.glm <- glm(Acceptance ~ GPA, data = MedGPA, family = binomial)
```

### Python

```{python}
#| label: python-fit-glm-gpa
import statsmodels.api as sm
import statsmodels.formula.api as smf

med_glm = smf.glm('Acceptance ~ GPA', data=MedGPA, family=sm.families.Binomial()).fit()
```

:::

The  argument `family=binomial` specifies that `Acceptance` follows a binomial distribution, with probability of success $p_i$ (i.e. the probability of the $i$th applicant being accepted is $p_i$). In addition, the probability $p_i$ is a function of $x_i$ (i.e. the probability of the $i$th applicant being accepted is a function of that applicant's GPA). 

The default [link function]{.term}, used is $\log\left(\dfrac{p_i}{1-p_i}\right)$. This function is known as the [logit function]{.term}. You can explicitly request it (unnecessarily) in R by writing `family = binomial(link="logit")` instead of `family = binomial`.

The model fit is shown below.

:::{.panel-tabset}

### R

```{r, echo=TRUE}
#| label: r-med-glm-summary
summary(med.glm)
```

### Python

```{python}
#| label: python-med-glm-summary2
print(med_glm.summary2())
```

:::

The regression equation for the fitted model is
$$\log \left(\dfrac{\hat{p}_i}{1-\hat{p}_i}\right)=-19.207+5.454 x_i,$$

or equivalently

$$\hat{p}_i=\dfrac{\exp(-19.207+5.454 x_i)}{1+\exp(-19.207+5.454 x_i)}.$$

The fitted curve for the probability of acceptance is shown below (in orange).

```{r, echo=FALSE, fig.height=4.5, message=FALSE}
#| label: r-medgpa-acceptance-fitted-curve
#| fig-alt: "Same scatterplot of GPA against acceptance with best fit line. Also a logistic curve in elongated S-shape low on left, high on right. Closer fit to points which remain all at Acceptance=0 or Acceptance=1."
medgpa.plot +
  geom_smooth(method = "lm",se = FALSE, fullrange = TRUE, colour = "#984ea3") +
  geom_smooth(method = "glm", color = "#ff7f00",se = FALSE, fullrange = TRUE,
              method.args = list(family = "binomial"))
```


We can see that that this curve fits the data better than the linear regression line, and that it gives predicted probabilities between 0 and 1, as desired. We could add predictors to the model to improve predictive performance -- we'll see more about that later.

The regression equation we have obtained allows us to predict the acceptance probability for a given GPA.

:::
<!-- end of Example -->

:::{.Task}
Predict the acceptance probability for an applicant with a GPA of (i) 2.5, (ii) 3 (iii) 4. First do this "by hand" using the regression equation, then in R using the `predict()` function.

*Hint: The* `predict()` *function will return values on the linear predictor scale unless you specify* `type='response'` *which returns probabilities instead.*
:::
<!-- end of Task -->

::::{.Answer}
In the *GPA and admission to medical school* example, we can write down the fitted model equation from the `summary(med.glm)`:
$$\log\left(\frac{p_i}{1-p_i}\right) =-19.207 + 5.454 \times \text{GPA}$$

From the fitted equation, we can obtain the acceptance probability by solving for $p_i$:

$$\hat{p}_i = \frac{\exp(-19.207 + 5.454 \times \text{GPA})} {1+ \exp(-19.207 + 5.454 \times \text{GPA})} $$

To predict the acceptance probability for an applicant we just need to substitute the specified GPA in the equation for $\hat{p}_i$:

(i) GPA = 2.5:
$$\hat{p}_i = \frac{\exp(-19.207 + 5.454 \times 2.5)} {1+ \exp(-19.207 + 5.454 \times 2.5)} \Rightarrow \hat{p}_i = 0.00378 $$

(ii) GPA = 3:
$$\hat{p}_i = \frac{\exp(-19.207 + 5.454 \times 3)} {1+ \exp(-19.207 + 5.454 \times 3)} \Rightarrow \hat{p}_i = 0.05494$$

(iii) GPA = 4:
$$\hat{p}_i = \frac{\exp(-19.207 + 5.454 \times 4)} {1+ \exp(-19.207 + 5.454 \times 4)} \Rightarrow \hat{p}_i = 0.93143$$


Alternatively, we can use the `predict()` function in `R` as follows:

```{r}
#| label: r-medgpa-predict-values
predict(med.glm, data.frame(GPA = c(2.5, 3, 4)), type = 'response')
```

::::
<!-- end of Answer -->

:::{.Weblink}
<https://goo.gl/mZHxyN>


 - Section 2.3 from *Mixed effects models and extensions in ecology with R -Zuur et al.* discusses the appropriateness of the assumptions of the linear model. 

:::
<!-- end Weblink -->


## What do the Bollywood box office and medical school admission examples have in common?

In both cases we have independent observations and we want to predict an outcome of interest (gross revenue/acceptance) based on an explanatory variable (budget/GPA). 

In both cases we have a regression equation allowing us to predict the response from a given value of the predictor. However, in one case the response is assumed to follow the normal distribution, in the other the binomial distribution. 

In both cases we fit a model to the *mean* of the response:

  * in the normal linear model the mean $E(Y_i)=\mu_i$ is assumed to be a linear function of $x_i$: $\mu_i=\beta_0+\beta_1x_i$, and 
  * in the logistic regression model the mean $\mu_i=E(Y_i)=p_i$ is modelled through the *logit link function*. That is, in logistic regression $\log\left(\dfrac{p_i}{1-p_i}\right)=\beta_0+\beta_1x_i$.
  
In slightly more general notation we can write $g(\mu_i)=\boldsymbol{x}_i^{\intercal}\boldsymbol{\beta}$ to cover both cases, where $\mu_i=E(Y_i)$ and $g(\mu_i)$ for each distribution is given in the following table.

| Model | Random component   | Systematic component   | Link function      |
|-------|--------------------|------------------------|--------------------|
| Normal model            | $y_i \overset{\text{indep}}{\sim} N(\mu_i, \sigma^2),$ <br> $E(Y_i) = \mu_i$ | $\boldsymbol{x}_i^{\intercal} \boldsymbol{\beta} = \beta_0 + \beta_1 x_i$ | Identity link: $g(\mu_i) = \mu_i$                                                          |
| Logistic regression model | $y_i \overset{\text{indep}}{\sim} Bin(1, p_i),$ <br> $E(Y_i) = p_i$ | $\boldsymbol{x}_i^{\intercal} \boldsymbol{\beta} = \beta_0 + \beta_1 x_i$ | Logit link: $g(\mu_i) = \log \left(\frac{\mu_i}{1-\mu_i}\right) = \log \left(\frac{p_i}{1-p_i}\right)$ |

: Summary table {tbl-colwidths="[15,30,25,30]"}


## Exponential family of distributions

As you may already know, there is a framework for classifying distributions, inside which both the Normal (Gaussian) and Binomial distributions can be seen as specific examples from the larger [exponential family]{.term}.

So are the Poisson, the Negative Binomial, Gamma distribution, Bernoulli (covered by Binomial already) and many others.

:::{.Definition}

## Exponential family of distributions

Consider a random variable $Y$ whose probability density function (p.d.f.) or probability mass function (p.m.f.) depends on parameter $\theta$. The distribution belongs to the exponential family if it can be written as 
$$
f(y; \theta) = \exp \Big[a(y) b(\theta)+c(\theta) + d(y) \Big].
$$
The term $b(\theta)$ is called the [natural parameter]{.term}. If $a(y)=y$ the distribution is said to be in [canonical form]{.term}.
:::
<!-- end of Definition -->


:::{.Example}

## Normal distribution is a member of exponential family

Consider $Y \sim N(\theta, \sigma^2)$ with p.d.f. $$
\begin{aligned} f(y; \theta) = \frac{1}{\sqrt{2 \pi \sigma^2}}\exp \left[ -\frac{1}{2\sigma^2} (y-\theta)^2\right], \hspace{0.5cm} -\infty <y<\infty.\end{aligned}
$$
 If we are interested in estimating $\theta$, the variance, $\sigma^2$, can be regarded as a nuisance parameter (which we may need to estimate at some other point). 
 
 Using the fact that $\blacksquare = \exp\left(\log(\blacksquare)\right)$, we rewrite the p.d.f. as $$
\begin{aligned} f(y; \theta) = \exp \left[-\dfrac{y^2}{2\sigma^2}+\dfrac{y \theta}{\sigma^2} -\dfrac{\theta^2}{2\sigma^2}-\dfrac{1}{2}\log(2 \pi \sigma^2)\right]\end{aligned}
$$
 we can see that this is of exponential family form with
 
| $a(y)$    | $b(\theta)$     | $c(\theta)$      | $d(y)$   |
|:--------:|:--------:|:-------------:|:----------------------:|
| $y$| $\frac{\theta}{\sigma^2}$ | $-\frac{\theta^2}{2\sigma^2}-\frac{1}{2}\log(2 \pi \sigma^2)$ | $-\frac{y^2}{2\sigma^2}$ |

: Normal distribution exponential family identification (for the case of fixed variance) {#tbl-normal-exp-family tbl-colwidths="[25,25,25,25]"}

So $a(y)=y$ meaning we can say it's in [canonical form]{.term}, and our [natural parameter]{.term} is $b(\theta)=\frac{\theta}{\sigma^2}$.

:::
<!-- end of Example -->


:::{.Task}

Show that the binomial distribution $\text{Bin}(n,p)$ is a member of the exponential family.
:::
<!-- end of Task -->

::::{.Answer}
Consider $Y \sim \text{Bin}(n,\theta)$ with p.m.f. $$
\begin{aligned} f(y; \theta) = \binom{n}{y} \theta^y (1-\theta)^{n-y}  \text{for } y=0,1,\dots,n. \end{aligned}
$$
 With extensive use of the same $\blacksquare = \exp\left(\log(\blacksquare)\right)$ trick, plus the log laws, we rewrite the p.m.f. as
 $$
\begin{aligned} f(y; \theta) &= \exp \left[ y \log \theta - y \log (1-\theta) + n \log(1-\theta) +\log \binom{n}{y} \right] \\ &= \exp \left[ y \log \frac {\theta}{1-\theta} + n \log(1-\theta) +\log \binom{n}{y} \right]  \end{aligned}
$$
 

we see that this is a member of the exponential family in canonical form, and its natural parameter is given by:
 $b(\theta)= \log \left(\dfrac{\theta}{1-\theta} \right)$.
::::
<!-- end of Answer -->

One of the reasons why seeing distrubutions as members of the exponential family is useful is that certain expectations and variances can be calculated for all such distributions simultaneously, using the general format.

In particular,

$$
E[a(Y)] = -\dfrac{c'(\theta)}{b'(\theta)}
$$
and
$$
\mathrm{Var}[a(Y)] =\dfrac{b''(\theta)c'(\theta)-c''(\theta)b'(\theta)}{[b'(\theta)]^3}.
$$

These may prove useful at some point, along with further useful properties of exponential family distributions related to the [score function]{.term} (we provide the general definition first).

:::{.Definition #rdefscoreuniv}
## Score statistic
Writing $l(\theta; y)$ for the log-likelihood, we define the [score statistic]{.term} $U$, by

$$
U\equiv \diff{l(\theta;y)}{\theta},
$$
i.e. as the derivative of the log-likelihood $l(\theta;y)$ with respect to the parameter $\theta$.
:::
<!-- end definition -->

Recall, that for exponential family distributions the log-likelihood is always just
$$
l(\theta; y) = a(y)b(\theta)+c(\theta) + d(y).
$$
and so the [score statistic]{.term} is simply
$$
U(\theta;y)=\diff{l(\theta;y)}{\theta}=a(y)b'(\theta)+c'(\theta)
$$


<!-- Remember that in *Learning from Data* you used  -->

The definition of the [score function]{.term} gives a new descriptive way to approach solving the maximum likelihood estimation problem. If we want to maximize the likelihood, then we typically just solve for the derivative being zero. So the [maximum likelihood estimate]{.term} $\hat{\theta}$ for any distributions in the exponential family can be found by solving
$$
U\equiv \diff{l(\theta;y)}{\theta} = 0.
$$

We can also think of the score statistic, $U=a(Y) b'(\theta)+c'(\theta)$, as a random variable in its own right, which means we can calculate its expectation
$$
\mathrm{E}(U)= b'(\theta)E[a(Y)]+c'(\theta)=b'(\theta)\left[-\frac{c'(\theta)}{b'(\theta)} \right]+c'(\theta)=0,
$$ 
which we note is always zero; and its variance 
$$
\mathrm{Var}(U)=[b'(\theta)^2]\mathrm{Var}[a(Y)].
$$

This leads us to a very important concept in statistical inference, called [Fisher information]{.term}, which we can use to obtain standard errors for maximum likelihood estimates of GLM coefficients.

:::{.Definition #rdefinfuniv}
## Fisher's information

The [Fisher Information]{.term}, denoted as $\mathcal{I}$, is given by:
$$
\mathcal{I} \equiv \mathrm{Var}(U)=E(U^2)=E\left[\left(\diff{l(\theta;y)}{\theta}\right)^2\right]=E\left[\diff[2]{l(\theta;y)}{{\theta}}\right].
$$

:::
<!-- end definition -->

<!--###[/definition]-->

The variance of the maximum likelihood estimates tells us about the amount of *information* that an observed random variable carries about an unknown parameter in the model that is linked to a distribution.

As we will see shortly, the score and information play a key role in parameter estimation and in obtaining standard errors for the coefficient estimates of a GLM.

Having defined the exponential family of distributions, we are now ready to formally define a GLM.


:::{.Weblink}
<https://goo.gl/mZHxyN>


 - Chapter 8 from *Mixed effects models and extensions in ecology with R -Zuur et al.* contains a more in-depth discussion of the exponential family. 

:::
<!-- end Weblink -->


## Generalised Linear Models

### The definition

:::{.Definition}

## Generalised Linear Models

Let $Y_i$ be independent responses from an exponential family distribution in canonical form and $\mu_i=E(Y_i)$, $i=1, \dots, n$.

A [generalised linear model]{.term} (GLM) is a model of the form
$$
g(\mu_i) = \boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}
$$ {#eq-glm-sys}

where $\boldsymbol{\beta}$ is a $p$-dimensional parameter vector,  $\boldsymbol{x}_i^{\intercal}$ is the $i$th row of the design matrix $\boldsymbol{X}$, and $g()$ is a monotonic, differentiable function called the [link function]{.term}.
:::
<!-- end of Definition -->

A GLM generalises the normal linear model by allowing:

  * a response variable with a distribution other than normal, but still a member of the exponential family of distributions; and

  * a non-trivial relationship between the response and the linear component of the form $g(\mu_i) = \boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}$ where $g()$ is the [link function]{.term}.

Importantly, $g(m)$ doesn't need to be $g(m)=m$, which it was in our ordinarly linear models.

### The [random component]{.term}

Suppose $Y_1,\dots, Y_n$ are independent random variables which follow an exponential family distribution such that $f(y_i; \theta_i)=\exp[y_i b(\theta_i)+c(\theta_i)+d(y_i)]$ for $i=1,\dots,n$.

The joint p.d.f. (or p.m.f.) of the $Y_i$ is 
$$
f(y_1,\dots,Y_n; \theta_1,\dots, \theta_n) =
  \prod_{i=1}^n \exp[y _i b(\theta_i)+c(\theta_i)+d(y_i)]
$$
So,
$$
f(y_1,\dots,Y_n; \theta_1,\dots, \theta_n) = \exp \left[ \sum_{i=1}^n y_i b(\theta_i)+ \sum_{i=1}^n c(\theta_i) +\sum_{i=1}^n d(y_i) \right]
$$ {#eq-joint}

The distribution of each $Y_i$ is in canonical form and depends on a single parameter $\theta_i$.

### The [systematic component]{.term}

Associated with each $y_i$ is a vector $\boldsymbol{x}_i=(x_{i1},x_{i2},\dots,x_{ip})^\intercal$ of values of $p$ explanatory variables. The response, $Y_i$, depends on the explanatory variables through a linear component,
$$
\eta_i=\boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}=\beta_1x_{i1}+\dots + \beta_p x_{ip}
$$
for $i=1,\dots, n.$

We will use the [link function]{.term} below to transform between the $y_i$ and the $\eta_i$ (though in some case we may use $y_i=\eta_i$), but the choice to use the values of the $p$ covariates to determing the $\eta$ is the [systematic component]{.term}.

The $i$th row of our design matrix, $\boldsymbol{X}$, is $\boldsymbol{x}_i^{\intercal}$; and $\boldsymbol{\beta}=(\beta_1,\dots, \beta_p)^\intercal$ is the parameter vector. Thus, as in linear models, the design matrix is given by $$\boldsymbol{X} = \left[ \begin{array}{c} \boldsymbol{x}_1^\intercal \\ \vdots \\ \boldsymbol{x}_n^\intercal \end{array}\right] =  \left[ \begin{array}{ccc} x_{11} & \dots & x_{1p} \\ \vdots & & \vdots \\ x_{n1} & \dots & x_{np}  \end{array}\right].$$

### The [link function]{.term}

The parameters $\theta_i$ in @eq-joint are usually not of direct interest as they are not available to us. Instead, we are interested in a smaller set of parameters $(\beta_1,\dots, \beta_p)$, and assume that $Y_i$ depends on these through the linear predictor $\eta_i$. The link between the distribution of the $Y_i$ and the linear predictor $\eta_i$ is provided by the link function $g()$, for which $g(\mu_i)=\eta_i=\boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}.$ Here $\mu_i=E(Y_i)$ and $g()$ is a monotone, differentiable function. Although any one-to-one function could be used in principle, certain choices of link function can offer great simplification. In particular, the link function can be chosen so that the natural parameter, $b(\theta_i)$, is proportional to the linear component $\eta_i=\boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}$. Such a link function is known as the [canonical link]{.term}. The following table shows the canonical link function for some of the most common distributions.

| Distribution | Natural parameter, $b(\theta)$              | Canonical link                                |
|-------------|------------------------------------------------|-----------------------------------------------|
| Normal      | $\dfrac{\theta}{\sigma^2}$                     | $g(\mu)=\mu$                                  |
| Poisson     | $\log \theta$                                  | $g(\mu)=\log(\mu)$                            |
| Binomial    | $\log\left(\dfrac{\theta}{1-\theta} \right)$  | $g(\mu)=\log\left(\dfrac{\mu}{1-\mu} \right)$ |

: {tbl-colwidths="[33,33,33]"}


:::{.Weblink}
<http://encore.lib.gla.ac.uk/iii/encore/record/C__Rb2939999?lang=eng>


 - Chapter 6 from *Extending linear models with R: generalized linear, mixed effects and nonparametric regression models - Faraway* gives an overview of GLMs and their properties.

:::
<!-- end Weblink -->


:::{.Weblink}
<https://goo.gl/mZHxyN>


 - Sections 9.1 and 9.2 from *Mixed effects models and extensions in ecology with R -Zuur et al.* cover the general formulation of GLMs. 

:::
<!-- end Weblink -->


Let us now look at some examples of generalised linear models and their components.


:::{.Example}

## Normal linear model

Consider $E(Y_i)=\mu_i =  \boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}$ where the $Y_i$ are independent  $N(\mu_i,\sigma^2)$  for $i=1,\dots,n$. Here $g(\mu_i)=\mu_i$, the [identity]{.term} link. You may be more familiar with this model written as $\boldsymbol{Y}=\boldsymbol{X}\boldsymbol{\beta}+\boldsymbol{\epsilon}$ where $\boldsymbol{\epsilon}=\left[ \begin{array}{c} \epsilon_1\\ \vdots \\ \epsilon_n  \end{array} \right]$ with the $\epsilon_i$ independent, identically distributed $N(0,\sigma^2)$ random variables.

So the standard linear model is an example of a generalised linear model. In much the same way a square is a rectangle.

:::
<!-- end of Example -->


<!-- Remove this example, its rather complex to understand

:::{.Example}

## A model for historical linguistics

If two languages are separated by time $t$, the probability of having cognate words for a particular meaning can be modelled as $\exp(-\theta t)$. For a test list of $n$ meanings, a linguist judges whether the corresponding words in two languages are cognate or not. For the $i$th meaning, define 
$$
Y_i = \left \lbrace \right.\begin{array}{ll} 1 & \text{if the two languages have cognate words} \\ 0 & \text{if words are not cognate.} \end{array}
$$
Then $P(Y_i=1)=\exp(-\theta t) =p$ and  $P(Y_i=0)=1-p$
i.e. $Y_i \sim \text{Bernoulli}(p)$ (or equivalently Bin$(1,p)$) and $E(Y_i)=p$. The link function is $g(p)= \log p = -\theta t$ so that $g(p)$ is linear in the parameter of interest, $\theta$; also $\boldsymbol{x}_i = -t$; and $\boldsymbol{\beta}=\theta$.
:::
<-- end of example -->

<!-- Removed cognate example above -->

:::{.Example}

Consider a model for whether students pass an end of semester exam at the first attempt. A Bernoulli variable $Y$ takes the value $1$ if they pass, and $0$ if not. Each student's performance contains some uncertainty, and is modelled by
$$
Y_i \sim \mathrm{Bernoulli}(p_i),
$$
where $p_i$ is their probability of passing.

Student $i$ attends a number of lectures, $l_i$, a number of labs $m_i$ and spends a time $t_i$ studying outside of classes.

We choose a model where these three covariates determine (through the [link function]{.term}) the probability of passing for each student, which as usual is the mean of $Y_i$.

Since we're working with a Bernoulli (or Binomial) our chosen linear model is that
$$
g(p_i) = \log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1l_i + \beta_2m_i+\beta_3t_i.
$$
So $g(p_i) = \boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}$ with $\boldsymbol{x}_i = \left(1, l_i,m_i, t_i\right)$ and $\boldsymbol{\beta} = \left(\beta_0, \beta_1, \beta_2, \beta_3\right)$.

:::
<!-- end of Example -->

:::{.Example}

## A model for mortality rates

 The number of deaths, $Y$, in a population in a year can be modelled by the Poisson distribution 
 $$
 f(y; \mu)=\frac{\mu^y e^{-\mu}}{y!}, \quad y=0,1,2,\dots.
 $$
 However the population itself changes in size over time, and different groups in the population have different average death rates. *Recall that a sum of Poissons is still Poisson*.
 
 So we can model the expected total number of deaths per year in a population as $E(Y)=\mu$ and it can be modelled by $$
E(Y)=\mu = n \lambda^*
$$
 where $n$ is the population size (in units of 100,000s) and $\lambda^*$ is the death rate per 100,000 people.
 
 To allow for different death rates by age, we can allow for the $\lambda^*$ to vary for different sub-populations.
 
 Let $Y_1,\dots, Y_n$ be the numbers of deaths occurring in successive age groups. A popular model we assume is that $E(Y_i)=\mu_i = n_i \exp(\theta \times i)$ where $Y_i \sim \mathrm{Poisson}(\mu_i)$ and
 
* $i=1$ for the age group 30-34, 
* $i=2$ for age group 35-39,
* $\vdots$
* $i=8$ for age group 65-69. 

So in this example $\lambda^*_i$ the death rate in sub-population $i$ is being written as $e^{\theta \times i}$, and $n_i$ denotes the number of people in that $i$th group (in our chosen units).

This model can be identified as a generalised linear model if we choose a log link function, i.e. $g(\mu_i) = \log \mu_i = \log n_i +\theta i$. Then we see that $g(\mu_i)$ our function of the distribution mean can be seen as a linear combination of covariates $\boldsymbol{x_i}$ and parameters $\boldsymbol{\beta}$ if we define them like this:
$$
\boldsymbol{x}_i^{\intercal}= (\log n_i, i) \text{ and } \boldsymbol{\beta}=(1, \theta)^{\intercal};
$$
so then
$$
g(\mu_i) = \boldsymbol{x}^{\intercal} \boldsymbol{\beta}.
$$
Each subpopulation comes with two covariates: $\log(n_i)$ the log of its population, and it's label $i$. The only parameter to estimate is $\theta$ as we know exactly how $\log(n_i)$ affects $\log(\mu_i)$ already by our model assumption.

When we see this idea later we will call the term $\log n_i$ is an [offset]{.term}, we will especially see more about it when we talk about models for counts.

:::
<!-- end of Example -->

:::{.Task}

Formulate the model used in the medical school admissions example as a GLM.
:::
<!-- end of Task -->

::::{.Answer}
Random component: Let $Y_i=1$ if the $i$th applicant is accepted to medical school and $Y_i=0$ if not. We assume that the $Y_i$ are independent responses from $\text{Bin}(1,p_i)$, with $E(Y_i)=p_i$ for $i=1,\dots,55$.

Systematic component: $\beta_0+\beta_1 x_i$ where $x_i$ is the $i$th applicant's GPA and $\beta_0$ and $\beta_1$ are parameters to be estimated.

Link function: $g(p_i)=\log \left(\dfrac{p_i}{1-p_i} \right)$ (logit link)

Equation of the GLM: $$\log \left(\dfrac{p_i}{1-p_i} \right)=\beta_0+\beta_1 x_i.$$ 
::::
<!-- end of Answer -->



<!-- # Summary  -->

<!-- Main concepts to be sure to be familiar with from this chapter: -->
<!-- \begin{itemize} -->
<!-- *  What are the two ways in which the GLM generalises the normal linear model? -->
<!-- *  What are the random and systematic component of a GLM and what is the role of the link function? -->
<!-- *  What is the form of the p.d.f. (or p.m.f.) of a distribution that is a member of the exponential family of distributions? What do the terms \textit{natural parameter}, \textit{canonical form} and \textit{canonical link} refer to? -->
<!-- *  Properties of exponential family distributions: can you derive/remember the expressions for the mean and variance of an exponential family distribution in canonical form? -->
<!-- \end{itemize} -->


## Maximum likelihood estimation of GLM coefficients

In a generalised linear model we are interested in the parameters $\beta_1,\dots,\beta_p$ that describe how the response depends on the explanatory variables. We use the observed $y_1,\dots,y_n$ to  maximise the log-likelihood function
$$
l(\boldsymbol{\beta},\boldsymbol{y})=\sum_{i=1}^n y_i b(\theta_i)+ \sum_{i=1}^n c (\theta_i) +\sum_{i=1}^n d(y_i)
$$ {#eq-loglik}

obtained from equation (@eq-joint). This depends on $\boldsymbol{\beta}$ through 

$$
\begin{aligned}
\mu_i & =E(Y_i)=-\frac{c'(\theta_i)}{b'(\theta_i)};    \\
\mathrm{Var}(Y_i) & = [b''(\theta_i)c'(\theta_i)-c''(\theta_i)b'(\theta_i)]/[b'(\theta_i)]^3;\\
g(\mu_i) & =\eta_i= \boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}, \quad i=1,\dots,n.
\end{aligned}
$$

The maximisation procedure results in $p$ simultaneous equations for $\hat{\boldsymbol{\beta}}$, which are usually solved numerically using the [method of scoring]{.term} (also known as [Fisher's scoring method]{.term}) and an algorithm called [iteratively reweighted least squares]{.term}.


<!-- To illustrate this process we will first present an example with a single parameter, $\theta$, of interest, before proceeding to develop the general theory of estimation of $(\beta_1,\dots,\beta_p)$ for generalised linear models. -->

<!-- \begin{example}[Weibull distribution for failure times] -->

<!-- Suppose we wish to obtain the maximum likelihood estimate of $\theta$ for independent random variables $Y_1, \dots, Y_n$ that follow the Weibull distribution with p.d.f. $$
\begin{aligned}f(y; \theta)= \exp[\log \lambda + (\lambda-1) \log y - \lambda \log \theta - (y/\theta)^\lambda]\end{aligned}
$$
 where $\lambda$ is assumed to be known. This distribution belongs to the exponential family with $$
\begin{aligned} &a(y)= y^\lambda, \nonumber\\ &b(\theta)=-\theta^{-\lambda}, \nonumber\\ &c(\theta)=\log \lambda - \lambda \log \theta,  \text{ and} \nonumber\\ &d(y)=(\lambda-1) \log(y). \hfill \text{[label: eqn:weibullexpfam]} \end{aligned}
$$
 The joint p.d.f. of the $Y_i$ is $$
\begin{aligned}f(y_1,\dots, y_n; \theta)= \prod_{i=1}^n \frac{\lambda y_i^{\lambda-1}}{\theta^\lambda} \exp\left[ - \left(\frac{y_i}{\theta} \right)^\lambda \right],\end{aligned}
$$
 and the corresponding log-likelihood is $l(\theta; y_1,\dots, y_n)= \sum_{i=1}^n \left[ [(\lambda-1) \log y_i + \log \lambda - \lambda \log \theta]- \left(\frac{y_i}{\theta} \right)^\lambda \right]. $ -->


<!-- The score function is the derivative of the log-likelihood with respect to $\theta$: $$
\begin{aligned} U(\theta)=\frac{dl}{d\theta}= \sum_{i=1}^n  \left[ -\frac{\lambda}{\theta}+\frac{\lambda y_i^\lambda}{\theta^{\lambda+1}} \right]. \hfill \text{[label: eqn:u]} \end{aligned}
$$
 The maximum likelihood estimator, $\hat{\theta}$, is the solution of the equation $U(\theta)=0 \Rightarrow \sum_{i=1}^n  \left[ -\frac{\lambda}{\theta}+\frac{\lambda y_i^\lambda}{\theta^{\lambda+1}} \right]=0$ Although this equation can be easily solved for $\theta$ (assuming $\lambda$ is known -- you can verify that the solution is $\left(\frac{1}{n} \sum y_i^{\lambda}\right)^{1/\lambda}$, for illustration purposes consider solving it numerically using the Newton-Raphson algorithm. The iterative process uses the formula $$
\begin{aligned} \theta^{(m)}= \theta^{(m-1)}-\frac{U^{(m-1)}}{U'^{(m-1)}}\hfill \text{[label: eqn:newtonraphson]} \end{aligned}
$$
 starting with an initial guess $\theta^{(1)}$ and obtaining successive approximations until convergence. The derivative, $U'$, with respect to $\theta$ is  $$
\begin{aligned} U'(\theta)= \frac{dU}{d \theta}= \sum_{i=1}^n \left[ \frac{\lambda}{\theta^2}-\frac{\lambda(\lambda+1)y_i^\lambda}{\theta^{\lambda+2}} \right] = \frac{\lambda n}{\theta^2}-\frac{\lambda(\lambda+1)\sum y_i^\lambda}{\theta^{\lambda+2}} \hfill \text{[label: eqn:uprime]} .\end{aligned}
$$
  Given a starting value, $\theta^{(1)}$, as well as $n$, $\sum y_i^\lambda$ and the expressions for $U$ and $U'$ in (*{ref{eqn:u}*) and (*{ref{eqn:uprime}*), we can use the Newton-Raphson formula (*{ref{eqn:newtonraphson}*) to obtain the maximum likelihood estimate of $\theta$. An alternative approach, known as the \term{method of scoring} or \term{Fisher scoring}, is to approximate $U'$ by its expectation $E(U')$ which, for exponential family distributions, is (see equation (*{ref{eqn:informationtheta}*))$E(U')=-\mathrm{Var}(U)=-\mathcal{I}=b''(\theta)\left[ -\frac{c'(\theta)}{b'(\theta)}\right]+c''(\theta) .$ Using the method of scoring we write the estimating equation as $$
\begin{aligned} \theta^{(m)}= \theta^{(m-1)}+\frac{U^{(m-1)}}{\mathcal{I}^{(m-1)}}. \hfill \text{[label: eqn:scoring]} \end{aligned}
$$
 In our example the information $\mathcal{I}$ is $$
\begin{aligned}\mathcal{I}=E(-U')= E\left[- \sum_{i=1}^n U'_i \right]=\sum_{i=1}^n E(-U'_i) =\sum_{i=1}^n \left[ \frac{b''(\theta)c'(\theta)}{b'(\theta)} -c''(\theta)\right]=\dots=\frac{\lambda^2 n}{\theta^2}, \hfill \text{[label: eqn:weibull]} \end{aligned}
$$
 -->
<!--   %\vspace{5cm} -->
<!--   where $U_i$ is the score for $Y_i$ and $b(\theta)$ and $c(\theta)$ are given in (*{ref{eqn:weibullexpfam}*). -->
<!-- \end{example} -->


<!--###[supplement] -->

:::{.Supplement}

Here we present in some detail how the maximum likelihood estimates of the coefficients of a GLM are obtained. Suppose that we are interested in estimating the parameter vector $(\beta_1,\dots, \beta_p)^\intercal$ in a GLM. To find the maximum likelihood estimates $\hat{\beta}_j$ we need the scores (multivariate version of the score from \longref{rdefscoreuniv}) expressed as functions of the $\beta_j$: 

$$
U_j\equiv \frac{\partial l}{\partial \beta_j} = \sum_{i=1}^n \left [   \frac{\partial l_i}{\partial \beta_j}  \right]= \sum_{i=1}^n \left [   \frac{\partial l_i}{\partial \theta_i} \cdot \frac{\partial \theta_i}{\partial \mu_i} \cdot \frac{\partial \mu_i} {\partial \beta_j}  \right]
$$ {#eq-chainrule}

For an exponential family distribution in canonical form, the components of @eq-chainrule are:

$$
\frac{\partial l_i}{\partial \theta_i} = y_i b'(\theta) +c'(\theta)=b'(\theta) \left[y_i -\left(-\frac{c'(\theta)}{b'(\theta)}\right) \right]=b' (\theta) (y_i - \mu_i) 
$$ {#eq-dldtheta}
$$
\begin{aligned}
& \frac{\partial \mu_i}{\partial \theta_i} = -\frac{c''(\theta_i)}{b'(\theta_i)}+\frac{c'(\theta_i) b''(\theta_i)}{[b'(\theta_i)]^2} = b'(\theta_i) \mathrm{Var}(Y_i)  \nonumber \\
\end{aligned}
$$
$$
\Rightarrow  \frac{\partial \theta_i}{\partial \mu_i}= \frac{1}{b'(\theta_i) \mathrm{Var}(Y_i)} 
$$ {#eq-dmudtheta}

$$
\frac{\partial \mu_i}{\partial \beta_j}=\frac{\partial \mu_i}{\partial \eta_i}\cdot \frac{\partial \eta_i}{\partial \beta_j}=\frac{\partial \mu_i}{\partial \eta_i}x_{ij}=\frac{x_{ij}}{g'(\mu_i)}
$$ {#eq-dmudbeta}


Substituting @eq-dldtheta, @eq-dmudtheta and @eq-dmudbeta into @eq-chainrule the expression for the scores becomes
 

$$
U_j=\sum_{i=1}^n \left[  \frac{(y_i-\mu_i)}{\mathrm{Var}(Y_i)}x_{ij} \frac{\partial \mu_i}{\partial \eta_i}\right]=\sum_{i=1}^n \left[  \frac{(y_i-\mu_i)}{\mathrm{Var}(Y_i)}\frac{x_{ij}} {g'(\mu_i)}\right].
$$ {#eq-score}


Note that the scores depend on $\boldsymbol{\beta}$ through $\mu_i=E(Y_i)$ and through $\mathrm{Var}(Y_i)$.
The variance-covariance matrix of the $U_j$ has terms $\mathcal{\mathcal{I}}_{jk}=E(U_jU_k)$ and is known as the [(Fisher) information matrix]{.term}. This is the multivariate version of \longref{rdefinfuniv}. The elements of matrix $\mathcal{\mathcal{I}}$ can be obtained from @eq-score:

$$
\begin{aligned}
 \mathcal{\mathcal{I}}_{jk}&=  E \left \lbrace \sum_{i=1}^n \left[  \frac{(y_i-\mu_i)}{\mathrm{Var}(Y_i)}x_{ij} \frac{\partial \mu_i}{\partial \eta_i}\right] \sum_{l=1}^n \left[  \frac{(y_l-\mu_l)}{\mathrm{Var}(Y_l)}x_{lk} \frac{\partial \mu_l}{\partial \eta_l}\right]  \right \rbrace \nonumber \\
 &=\sum_{i=1}^n \frac{E[(Y_i-\mu_i)^2]x_{ij}x_{ik}}{[\mathrm{Var}(Y_i)]^2} \left(\frac{\partial \mu_i}{\partial \eta_i} \right)^2  \nonumber\\
\end{aligned}
$$
$$
=\sum_{i=1}^n \frac{x_{ij}x_{ik}}{\mathrm{Var}(Y_i)} \left(\frac{\partial \mu_i}{\partial \eta_i} \right)^2 = \sum_{i=1}^n \frac{x_{ij}x_{ik}}{\mathrm{Var}(Y_i)\left(g'(\mu_i)\right)^2}
$$ {#eq-informationmatrix}


Here we have used the fact that $E[(Y_i-\mu_i)(Y_l-\mu_l)]=0$ by the independence of the $Y_i$.
Notice that the information matrix can be written as 

$$
\mathcal{I}=\mathcal{I}(\boldsymbol{\beta})=\boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{X}
$$ {#eq-informationirwls}

 where $\boldsymbol{X}=\left[ \begin{array}{c} \boldsymbol{x}_1^\intercal\\ \dots \\  \boldsymbol{x}_n^\intercal \end{array} \right]= \left[ \begin{array}{ccc}
x_{11} & \dots & x_{1p}\\
\vdots & \ddots & \vdots\\
x_{n1} & \dots & x_{np}                                                                                                                                                                                                                     \end{array} \right],$
 $\boldsymbol{W}=\text{diag}(\boldsymbol{w})=  \left[ \begin{array}{cccc}
w_1 & 0 &  \dots & 0\\
0 & w_2 & & \vdots\\
\vdots & & \ddots & 0\\
0 & \dots &0 & w_n                                                                                                                                                                                                                     \end{array} \right],$ and 
$$
\begin{aligned}
w_i=\frac{1}{\mathrm{Var}(Y_i)\left(g'(\mu_i)\right)^2}, \hspace{1cm} i=1,\dots, n.
\end{aligned}
$$

The information matrix $\mathcal{I}(\boldsymbol{\beta})$ depends on $\boldsymbol{\beta}$ through $\boldsymbol{\mu}$ and through $\mathrm{Var}(Y_i)$ for $i=1,\dots,n$.

@eq-score can be written as 

$$
U_j=\sum_{i=1}^n (y_i-\mu_i)x_{ij}w_i g'(\mu_i)=   \sum_{i=1}^n x_{ij}w_iz_i     \hspace{1cm} j=1,\dots,p
$$ {#eq-scoreirwls}



where $z_i=(y_i-\mu_i)g'(\mu_i)$, so the score can be expressed in vector-matrix form as 

$$
\boldsymbol{U}(\boldsymbol{\theta})= \boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{z} .
$$ {#eq-scoreirwlsmatrix}




Fisher's method of scoring is based on the estimating equation 

$$
\hat{\boldsymbol{\beta}}^{(m)} = \hat{\boldsymbol{\beta}}^{(m-1)} +[\mathcal{\mathcal{I}}^{(m-1)}]^{-1} \boldsymbol{U}^{(m-1)}
$$ {#eq-scoringpdim}

 where $\hat{\boldsymbol{\beta}}^{(m)}$ is the vector of estimates of $(\beta_1,\dots \beta_p)$ at the $m$th iteration, $[\mathcal{\mathcal{I}}^{(m-1)}]^{-1}$ is the inverse of the information matrix with elements $\mathcal{\mathcal{I}}_{jk}$ given by @eq-informationmatrix, and $\boldsymbol{U}^{(m-1)}$ is the vector of elements given by @eq-score, all evaluated at  $\hat{\boldsymbol{\beta}}^{(m-1)}$.
  Substituting @eq-informationirwls and @eq-scoreirwlsmatrix in @eq-scoringpdim we obtain 
$$
\begin{aligned}
\hat{\boldsymbol{\beta}}^{(m)}&=   \hat{\boldsymbol{\beta}}^{(m-1)}+\left[\boldsymbol{X}^\intercal \boldsymbol{W}^{(m-1)} \boldsymbol{X}\right]^{-1}  \boldsymbol{X}^\intercal \boldsymbol{W}^{(m-1)} \boldsymbol{z}^{(m-1)}\\
  &=  \left[\boldsymbol{X}^\intercal \boldsymbol{W}^{(m-1)} \boldsymbol{X}\right]^{-1}  \boldsymbol{X}^\intercal \boldsymbol{W}^{(m-1)}\boldsymbol{X} \hat{\boldsymbol{\beta}}^{(m-1)}+\left[\boldsymbol{X}^\intercal \boldsymbol{W}^{(m-1)} \boldsymbol{X}\right]^{-1}  \boldsymbol{X}^\intercal \boldsymbol{W}^{(m-1)} \boldsymbol{z}^{(m-1)},
\end{aligned}
$$
this simplifies slightly to
$$
\hat{\boldsymbol{\beta}}^{(m)} =\left[\boldsymbol{X}^\intercal \boldsymbol{W}^{(m-1)} \boldsymbol{X}\right]^{-1}  \boldsymbol{X}^\intercal \boldsymbol{W}^{(m-1)} (\boldsymbol{\eta}^{(m-1)}+\boldsymbol{z}^{(m-1)})
$$ {#eq-irwls}

 This is the same form as the normal equations for weighted least squares, except that it has to be solved numerically because $\boldsymbol{\eta}$, $\boldsymbol{z}$ and $\boldsymbol{W}$ depend on $\hat{\boldsymbol{\beta}}$.
 
This is why the method to obtain maximum likelihood estimators for GLMs is called [iteratively (re)weighted least squares (IRWLS)]{.term}.
The procedure begins by using an initial approximation $\hat{\boldsymbol{\beta}}^{(0)}$ to obtain estimates of $\boldsymbol{\eta}$, $\boldsymbol{z}$ and $\boldsymbol{W}$. Then $\hat{\boldsymbol{\beta}}^{(1)}$ is obtained from @eq-irwls and is used to update $\boldsymbol{\eta}$, $\boldsymbol{z}$ and $\boldsymbol{W}$. The iterative process continues until the difference between successive approximations $\hat{\boldsymbol{\beta}}^{(m-1)}$ and $\hat{\boldsymbol{\beta}}^{(m)}$ is sufficiently small.

:::
<!-- ###[/supplement] -->



<!-- \begin{example}[Poisson regression] -->

<!-- Suppose we have the following data: -->
<!-- \begin{center} -->
<!-- \begin{tabular}{crrrrrrrrrr} -->
<!-- \hline -->
<!-- $y_i$ & 2 & 3 & 6 & 7 & 8 & 9 & 10& 12 & 15\\ -->
<!-- $x_i$ & -1 & -1 & 0 & 0 &0 & 0 & 1 & 1 & 1 \\ -->
<!-- \hline -->
<!-- \end{tabular} -->
<!-- \end{center} -->
<!-- Assume that the responses $Y_i$ are independent Poisson random variables with $E(Y_i)=\mu_i = \beta_1 + \beta_2 x_i= \boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}=\eta_i$ where $\boldsymbol{\beta}= \left[ \begin{array}{c} \beta_1\\ \beta_2 \end{array}\right] \text{ and } \boldsymbol{x}_i =   \left[ \begin{array}{c} 1\\ x_i \end{array} \right] \text{ for } i=1,\dots, n.$ Note that here we are taking the link function to be the identity link: $g(\mu_i)= \mu_i= \boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}=\eta_i$ -->
<!-- Then $g'(\mu_i)=\dfrac{\partial \mu_i}{\partial \eta_i}= 1$ so the weights in (*{ref{eqn:informationirwls}*) are  $w_i = \frac{1}{\mathrm{Var}(Y_i)} =\frac{1}{\mu_i}=\frac{1}{\beta_1+\beta_2 x_i}$ and $\eta_i+z_i$ in  (*{ref{eqn:irwls}*) simplifies to  $\eta_i+z_i = -->
<!--  \hat{\beta}_1+\hat{\beta}_2x_i+ (y_i-\mu_i)=\hat{\beta}_1 +\hat{\beta}_2 x_i +(y_i - \hat{\beta}_1 -\hat{\beta}_2 x_i) = y_i. $ Also $\mathcal{\mathcal{I}}=\boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{X}= \left[ \begin{array}{cc} \sum \dfrac{1}{\hat{\beta}_1+\hat{\beta}_2 x_i }&  \sum \dfrac{x_i}{\hat{\beta}_1+\hat{\beta}_2 x_i} \\ -->
<!--  \sum \dfrac{x_i}{\hat{\beta}_1+\hat{\beta}_2 x_i}  &  \sum \dfrac{x_i^2}{\hat{\beta}_1+\hat{\beta}_2 x_i } \end{array}\right] $ -->
<!-- and $\boldsymbol{X}^\intercal \boldsymbol{W} (\boldsymbol{\eta}+\boldsymbol{z})= \left[ \begin{array}{c} \sum \dfrac{y_i}{\hat{\beta}_1+\hat{\beta}_2 x_i }\\ -->
<!--  \sum \dfrac{x_i y_i}{\hat{\beta}_1+\hat{\beta}_2 x_i}  \end{array}\right] $ -->
<!-- so that the MLEs can be obtained from  $\hat{\boldsymbol{\beta}}^{(m)}=\left[(\boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{X})^{(m-1)}\right]^{-1} (\boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{z}) ^{(m-1)}.$ -->
<!-- For the data in this example, $\boldsymbol{\eta}+ \boldsymbol{z}=\boldsymbol{y}=\left[\begin{array}{c} 2\\ 3\\ \vdots\\ 15 \end{array}\right] \text{ and } \boldsymbol{X}=\left[ \begin{array}{c} \boldsymbol{x}_1^\intercal \\ \boldsymbol{x}_2^\intercal \\ \vdots \\ \boldsymbol{x}_9^\intercal  \end{array} \right]=\left[\begin{array}{rr} 1 &  -1\\ 1 &  -1\\ \vdots & \vdots\\ 1 & 1 \end{array}\right]. $ -->
<!-- We also need some initial parameter estimates. Start with $\hat{\beta}_1^{(1)}=7$ and $\hat{\beta}_2^{(1)}=5$. -->
<!-- Then $(\boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{X})^{(1)}= \left[ \begin{array}{rr} 1.821429 & -0.75  \\ -->
<!--  -0.75&  1.25 \end{array}\right], $ -->
<!--  $(\boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{z})^{(1)}= \left[ \begin{array}{cc} 9.869048  \\ -->
<!-- 0.583333 \end{array}\right], $ -->
<!--  so $$
\begin{aligned} \hat{\boldsymbol{\beta}}^{(2)}= [(\boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{X})^{(1)}]^{-1}(\boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{z})^{(1)}= \left[ \begin{array}{cc} 7.4514  \\ -->
<!-- 4.9375 \end{array}\right] -->
<!-- \end{aligned}
$$
 -->
<!-- and so on until convergence.\\ -->
<!-- The MLEs are $\hat{\beta}_1=7.45163$ and $\hat{\beta}_2=4.93520$. -->
<!-- \end{example} -->

<!-- \begin{remark} Fisher scoring and the Newton-Raphson algorithm are equivalent when the canonical link is used in a generalised linear model. This is because for the canonical link function $\mathcal{I}=E(U')=-U'$. -->
<!-- \end{remark} -->

<!-- \begin{remark} The linear model is a generalised linear model with identity link, $\eta_i=g(\mu_i)=\mu_i$, and $\mathrm{Var}(Y_i)=\sigma^2$ for $i=1,\dots,n$. This results in weights $w_i=1/\sigma^2$ and $z_i=(y_i-\mu_i)$ that do not depend on $\boldsymbol{\beta}$. Hence $\boldsymbol{\eta}+\boldsymbol{z}=\boldsymbol{y}$ and $\boldsymbol{W}=1/\sigma^2 \boldsymbol{I}$, where $\boldsymbol{I}$ is the $n\times n$ identity matrix. Therefore the solution to the likelihood equations can be obtained in one step as we will see in more detail in Chapter 5. -->
<!-- \end{remark} -->

<!-- #Summary -->
<!-- Points to take away from this chapter: -->
<!-- \begin{itemize} -->
<!-- *  Multivariate versions of the score and information (score statistics and information matrix) -->
<!-- *  How to do maximum likelihood estimation -->
<!-- *  Finding the maximum likelihood estimator using numerical methods (Newton-Raphson method, Fisher's scoring method/iteratively reweighted least squares) -->
<!-- *  Think about estimation in the context of a GLM: if you are asked how the estimates for the model parameters were obtained, can you name the estimation method and briefly describe how it works? -->
<!-- \end{itemize} -->

## Inference for GLMs
We will now turn our attention to inference for GLMs, mainly through hypothesis tests and the construction of confidence intervals for the parameters of interest. For that we need some sampling distribution results.
<!-- We can think of this process in the context of two models which have the same probability distribution and the same link function, but different number of parameters in the linear component. The simpler model (model under the null hypothesis $H_0$) must be a special case of the more general model. Then hypothesis testing takes the steps: -->
<!--  \begin{enumerate} -->
<!--  *  Specify model $M_0$ to correspond to $H_0$ and a more general model $M_1$ with $M_0$ a special case of $M_1$. -->
<!--    *  Fit $M_0$ and calculate a ``goodness of fit statistic" $G_0$. Fit $M_1$ and the corresponding statistic $G_1$. -->
<!--    *   Calculate the improvement in fit, usually $G_1-G_0$ (or sometimes $G_1/G_0$). -->
<!--     *  Use the sampling distribution of $G_1-G_0$ to test the null hypothesis that $G_1=G_0$ against $G_1 \neq G_0$. -->
<!--         *  If the hypothesis $G_1=G_0$ is not rejected, then $H_0$ is not rejected and $M_0$ is the preferred model. If $G_1=G_0$ is rejected, then $M_1$ is regarded as the better model. -->
<!--  \end{enumerate} -->
<!-- To obtain confidence intervals and test statistics we need the sampling distribution of the estimator. Exact sampling distributions can be obtained if the response follows the normal distribution. Otherwise, we use asymptotic (large sample) results, usually based on the Central Limit Theorem. For these results to be valid, certain conditions need to hold, but we won't go into them in detail because these are satisfied for the exponential family distributions used in the GLMs we will be covering. -->

<!-- #Large sample approximations  -->

<!-- To come up with the approximate distribution of a statistic of interest, we rely on the following large sample distribution results: If $S$ is a statistic of interest, under appropriate conditions $\dfrac{S-E(S)}{\sqrt{\mathrm{Var}(S)}}  \overset{\text{approx}}{\sim} N(0,1)$, or equivalently  $\dfrac{[S-E(S)]^2}{\mathrm{Var}(S)}  \overset{\text{approx}}{\sim} \chi^2(1)$. -->
<!--  For a vector of statistics of interest $\boldsymbol{s}=(S_1, S_2, \dots, S_p)^\intercal$ with mean $E(\boldsymbol{s})$ and variance-covariance matrix $\boldsymbol{V}$, we have the asymptotic result -->
<!--  $$
\begin{aligned} [\boldsymbol{s}-E(\boldsymbol{s})]^\intercal \boldsymbol{V}^{-1}  [\boldsymbol{s}-E(\boldsymbol{s})]  \overset{\text{approx}}{\sim} \chi^2(p), \hfill \text{[label: eqn:approxnormgeneral]}\end{aligned}
$$
 provided that $\boldsymbol{V}$ is not singular. -->

<!-- #Sampling distribution for score statistics  -->
<!--  Suppose $Y_1, \dots, Y_n$ are independent random variables in a GLM with parameters $\boldsymbol{\beta}$, with $E(Y_i)=\mu_i$ and $g(\mu_i)=\boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}= \eta_i$. -->

<!-- Recall equation (*{ref{eqn:score}*), which gives the score statistics as $$
\begin{aligned}U_j= \frac{\partial l}{\partial \beta_j} = \sum_{i=1}^n \left[ \frac{(Y_i-\mu_i)}{\mathrm{Var}(Y_i)}\frac{x_{ij}}{g'(\mu_i)} \right]  \hspace{1cm}  \text{ for } j=1, \dots, p. \end{aligned}
$$
 -->
<!-- Recall also that since $E(Y_i)=\mu_i$, we have that $E(U_j)=0$ for $j=1,\dots, p$ and that the variance-covariance matrix of the score statistics has elements -->
<!-- $$
\begin{aligned}\mathcal{I}_{jk}= E(U_jU_k).\hfill \text{[label: eqn:scorevar]}\end{aligned}
$$
 -->

<!-- For univariate $\beta$, the score statistic has asymptotic sampling distribution $$
\begin{aligned}\frac{U}{\sqrt{\mathcal{I}}} \overset{\text{approx}}{\sim}N(0,1)\hfill \text{[label: eqn:scoreuniv]}\end{aligned}
$$
 or equivalently  $$
\begin{aligned}\frac{U^2}{\mathcal{I}} \overset{\text{approx}}{\sim} \chi^2(1).\end{aligned}
$$
 -->
<!-- For $\boldsymbol{\beta}=(\beta_1, \dots, \beta_p)^\intercal$, the score vector $$
\begin{aligned}\boldsymbol{U}=(U_1,\dots, U_p)^\intercal \overset{\text{approx}}{\sim} MVN(\boldsymbol{0}, \mathcal{I})  \end{aligned}
$$
 and so $$
\begin{aligned} \boldsymbol{U}^\intercal \mathcal{I}^{-1} \boldsymbol{U} \overset{\text{approx}}{\sim} \chi^2(p).\hfill \text{[label: eqn:scoremultiv]}\end{aligned}
$$
 -->

<!-- \begin{example}[Normal distribution] -->
<!-- Let $Y_1, \dots, Y_n$ be independent $N(\mu, \sigma^2)$ random variables where $\sigma^2$ is assumed to be known. The log-likelihood function is -->
<!--  $$
\begin{aligned}l = -\frac{1}{2\sigma^2}\sum_{i=1}^n(y_i -\mu)^2 -n \log(\sqrt{2\pi \sigma^2}).\hfill \text{[label: eqn:infnormal]} -->
<!--  \end{aligned}
$$
 -->
<!--  The score statistic is $$
\begin{aligned}U=\frac{dl}{d\mu}= \frac{1}{\sigma^2}\sum (Y_i -\mu) = \frac{n}{\sigma^2} (\bar{Y}-\mu).\hfill \text{[label: eqn:scorenormal]}\end{aligned}
$$
 -->
<!-- It can be seen that the expectation of $U$ is $0$ since $E(Y_i)=\mu$. -->
<!-- The variance of $U$ is -->
<!--  $$
\begin{aligned} \mathcal{I} = \mathrm{Var}(U)= \frac{1}{\sigma^4}\sum \mathrm{Var}(Y_i)=\frac{n}{\sigma^2}. \hfill \text{[label: eqn:varnormal]}\end{aligned}
$$
 -->
<!--  Therefore $$
\begin{aligned}\frac{U}{\sqrt{\mathcal{I}}}= \frac{\bar{Y}-\mu}{\sigma / \sqrt{n}} \overset{\text{approx}}{\sim}N(0,1).\hfill \text{[label: eqn:scoredistnormaluniv]} \end{aligned}
$$
 In fact, this is an exact result since $\bar{Y} \sim N(\mu, \sigma^2/n)$. -->
<!-- Also, $$
\begin{aligned} \boldsymbol{U}^\intercal \mathcal{I}^{-1} \boldsymbol{U}= \frac{U^2}{\mathcal{I}} = \frac{(\bar{Y}-\mu)^2}{\sigma^2/n} \sim \chi^2(1).\hfill \text{[label: eqn:scoredistnormalmultiv]}\end{aligned}
$$
 -->
<!-- This sampling distribution can be used obtain interval estimates for $\mu$ of the form $\bar{y} \pm 1.96 \sigma /\sqrt{n} $ where $\sigma^2$ is assumed to be known. -->
<!-- \end{example} -->

<!-- \begin{example}[Binomial distribution] -->
<!-- For $Y \sim Bin(n,p)$, the log-likelihood function is -->
<!--  $$
\begin{aligned}l(p; y) = y \log p+(n-y) \log(1-p)+ \log \binom{n}{y}.\hfill \text{[label: eqn:binomialloglik]}\end{aligned}
$$
 -->
<!--  The score statistic is $$
\begin{aligned}U=\frac{dl}{dp}= \frac{Y}{p}-\frac{n-Y}{1-p} = \frac{Y-np}{p(1-p)},\hfill \text{[label: eqn:scorebinomial]} -->
<!--  \end{aligned}
$$
 -->
<!--    and the expectation of $U$ is $0$ since $E(Y)=np$. -->
<!-- Also, $\mathrm{Var}(Y)=np(1-p)$, so $$
\begin{aligned}\mathcal{I}=\mathrm{Var}(U) = \frac{1}{p^2(1-p)^2} \mathrm{Var}(Y)=\frac{n}{p(1-p)}.\hfill \text{[label: eqn:infbinomial]}\end{aligned}
$$
 -->
<!-- Therefore a large sample approximation for $U$ is $$
\begin{aligned}\frac{U}{\sqrt{\mathcal{I}}} = \frac{Y-np}{\sqrt{np(1-p)}} \overset{\text{approx}}{\sim} N(0,1).\hfill \text{[label: eqn:scoredistbinom]}\end{aligned}
$$
 -->
<!-- This is the normal approximation to the binomial distribution (without a continuity correction). It can be used to obtain approximate confidence intervals for $p$. -->
<!-- \end{example} -->

### Sampling distribution of the MLE
<!-- To obtain an asymptotic distribution of the MLE in a generalised linear model, we need to consider some Taylor series expansions of the log-likelihood and its derivatives.\\ -->

<!-- \term{Reminder:} The \term{Taylor series expansion} of a function $f(x)$ about a value $t$ is given by $f(x)= f(t)+ (x-t) \left. \frac{df}{dx} \right|_{x=t}+ \frac{1}{2} (x-t)^2 \left. \frac{d^2f}{dx^2} \right|_{x=t} + \dots$ provided that $x$ is near $t$.\\ -->


<!-- Let's begin with an approximation for the log-likelihood: -->
<!-- Consider the first three terms of the expansion $$
\begin{aligned}l(\beta) & \approx l(\hat{\beta})+ (\beta-\hat{\beta}) \left. \frac{dl}{d\beta} \right|_{\beta=\hat{\beta}}+ \frac{1}{2} (\beta-\hat{\beta})^2 \left. \frac{d^2l}{d\beta^2} \right|_{\beta=\hat{\beta}} \\ & \approx l(\hat{\beta}) + (\beta-\hat{\beta}) U(\hat{\beta}) +  \frac{1}{2} (\beta-\hat{\beta})^2 U'(\hat{\beta})\end{aligned}
$$
 -->

<!-- Replacing $U'=d^2l/ d\beta^2$ by its expected value, $E(U')=-\mathcal{I}$, the approximation becomes  $l(\beta) \approx l(\hat{\beta}) + (\beta-\hat{\beta}) U(\hat{\beta}) -  \frac{1}{2} (\beta-\hat{\beta})^2 \mathcal{I}(\hat{\beta}).$ -->

<!--  For a vector $\boldsymbol{\beta}$ of parameters this generalizes to $$
\begin{aligned} l(\boldsymbol{\beta}) \approx l(\hat{\boldsymbol{\beta}}) + (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}})^\intercal \boldsymbol{U}(\hat{\boldsymbol{\beta}}) -  \frac{1}{2} (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}})^\intercal \mathcal{I}(\hat{\boldsymbol{\beta}})(\boldsymbol{\beta}-\hat{\boldsymbol{\beta}}). \hfill \text{[label: eqn:likelihoodexp]} \end{aligned}
$$
 -->

<!-- Similarly for the score function with a single parameter $\beta$ the Taylor series expansion is $$
\begin{aligned}U(\beta) & \approx U(\hat{\beta})+ (\beta-\hat{\beta})U'(\hat{\beta}).\end{aligned}
$$
 -->
<!-- Approximating $U'$ by its expectation $E(U')=-\mathcal{I}$ this becomes -->
<!-- $$
\begin{aligned}U(\beta) & \approx U(\hat{\beta})- (\beta-\hat{\beta})\mathcal{I}(\hat{\beta}).\end{aligned}
$$
 -->
<!-- For a vector $\boldsymbol{\beta}$ of parameters this generalizes to $$
\begin{aligned} U(\boldsymbol{\beta}) \approx U(\hat{\boldsymbol{\beta}}) -\mathcal{I}(\hat{\boldsymbol{\beta}}) (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}}).\hfill \text{[label: eqn:scoreexp]}\end{aligned}
$$
 -->

<!-- Equation (*{ref{eqn:scoreexp}*) can be used to obtain the sampling distribution of the MLE $\hat{\boldsymbol{\beta}}$.  Since $U(\hat{\boldsymbol{\beta}})=\boldsymbol{0}$, we have  $ U(\boldsymbol{\beta}) \approx  -\mathcal{I}(\hat{\boldsymbol{\beta}}) (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}}) \Rightarrow (\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}) = \mathcal{I}^{-1} \boldsymbol{U} $ assuming $\mathcal{I}$ is not singular. Taking the expectation (regarding $\mathcal{I}$ as a constant), we have that asymptotically $\hat{\boldsymbol{\beta}}- \boldsymbol{\beta}=\boldsymbol{0}$ since $E(\boldsymbol{U})=0$. Thus $E(\hat{\boldsymbol{\beta}})=\boldsymbol{\beta}$ (asymptotically) and hence the MLE is asymptotically unbiased. -->
<!--  The variance-covariance matrix for $\hat{\boldsymbol{\beta}}$ is $E\left[(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta})(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta})^\intercal  \right]= \mathcal{I}^{-1} E(\boldsymbol{U}\boldsymbol{U}^\intercal) \mathcal{I}^{-1} = \mathcal{I}^{-1}$ since $E(\boldsymbol{U}\boldsymbol{U}^\intercal)=\mathcal{I}$ and $(\mathcal{I}^{-1})^\intercal =\mathcal{I}^{-1}$ since $\mathcal{I}$ is symmetric. -->
<!-- Then by equation (*{ref{eqn:approxnormgeneral}*), -->




The asymptotic (large sample) distribution for $\hat{\boldsymbol{\beta}}$ is
  

$$
(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta})^\intercal \mathcal{I}(\hat{\boldsymbol{\beta}}) (\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}) \overset{\text{approx}}{\sim} \chi^2(p)
$$ {#eq-waldchi}

 or equivalently 

$$
\hat{\boldsymbol{\beta}} \overset{\text{approx}}{\sim} N_p(\boldsymbol{\beta}, \mathcal{I}^{-1}),
$$ {#eq-waldnormal}

 where $\chi^2(p)$ refers to the chi-squared distribution with $p$ degrees of freedom and $N_p(\boldsymbol{\beta},\mathcal{I}^{-1})$ refers to a multivariate ($p$-dimensional) normal distribution with vector mean $\boldsymbol{\beta}$ and $p \times p$-dimensional variance-covariance matrix $\mathcal{I}^{-1}$.

In the one-dimensional case, what this says is that the MLE $\hat{\beta}$ is approximately normally distributed with mean $\beta$ and variance $\mathcal{I}^{-1}$ (scalars).

This allows us to perform hypothesis tests and construct confidence intervals for the model parameters.


:::{.Definition}

## Wald statistic

The [Wald statistic]{.term}, also known as the [z-statistic]{.term}, for each of the model parameters $\left \lbrace \beta_j\right \rbrace,~j=1,\dots,p$, is equal to the coefficient estimate, $\hat{\beta}_j$, over its  standard error, $\text{se}(\hat{\beta}_j)$.
:::
<!-- end of Definition -->

Under the null hypothesis $H_0: \beta_j=0$, and using the asymptotic normality result for the MLE, the Wald statistic is approximately distributed as standard normal. In other words, we have that $$
z=\frac{\hat{\beta}_j}{\text{se}(\hat{\beta}_j)} \overset{\text{approx}}{\sim} N(0,1).
$$ This allows us to perform the [Wald test]{.term}, which compares the z-statistic to the extreme percentiles of a standard normal distribution.

Also using this asymptotic normality result, we can construct an approximate 95\% confidence interval for $\beta_j$ by taking $$
\hat{\beta}_j\pm 1.96\text{se}(\hat{\beta}_j).
$$ Here 1.96 is the 97.5th percentile of the standard normal distribution.


:::{.Example}

## Hypothesis test and confidence interval for the GPA coefficient in the model for medical school admissions

Recall the logistic regression model for the medical school admissions data. In the output we see the MLEs of $\beta_0$ and $\beta_1$ in the `Estimate` column. These are obtained by solving the likelihood equations numerically using Fisher's scoring method (notice the `Number of Fisher Scoring iterations` information towards the end.)

```{r echo=c(4:6)}
#| label: r-medgpa-hypoth-glm-summary
library(Stat2Data)
data(MedGPA)
med.glm <- glm(Acceptance ~ GPA, data=MedGPA, family=binomial)
summary(med.glm)
```

We can also see the standard errors for $\hat{\beta_0}$ `(Intercept)` and $\hat{\beta_1}$ (GPA coefficient). These are obtained from the observed information matrix, which is computed during the iterative estimation procedure. Remember that the variance of $\boldsymbol{\beta}$ is estimated by $\mathcal{I}^{-1}$, the inverse of the information matrix. The R function `vcov()`returns this estimated variance-covariance matrix of the model coefficients:

```{r}
#| label: r-medgpa-hypoth-vcov
vcov(med.glm)
```

The diagonal entries of this matrix are the estimated variances for $\hat{\beta_0}$ and $\hat{\beta_1}$ and the off-diagonal entries give the estimated covariance between $\hat{\beta_0}$ and $\hat{\beta_1}$. The standard errors shown in the output (`Std. Error` column) are the square roots of the estimated variances:

```{r}
#| label: r-medgpa-hypoth-sqrt-diag
sqrt(diag(vcov(med.glm)))
```

The `z value` column gives the Wald statistics that test the hypotheses $H_0: \beta_0=0$ and $H_0: \beta_1=0$. We are almost never interested in testing whether the intercept term is zero or not, we focus instead on the coefficients of the explanatory variables in the model. So for $H_0: \beta_1=0$ the $z$ value equals 3.454 and is obtained by taking the coefficient estimate and dividing by its standard error $$z=\frac{\hat{\beta_1}}{\text{se}(\hat{\beta}_1)}=\frac{5.454}{1.579}=3.454.$$ Under $H_0$, $z$ should approximately follow the standard normal distribution, $N(0,1)$. The $p$-value shown as `Pr(>|z|)` in the output is the probability of obtaining a value as extreme as $z$ or larger in absolute value, assuming the null hypothesis is true. A small $p$-value indicates that the $z$ value is unlikely to come from a $N(0,1)$ distribution and leads to rejecting $H_0$. As the $p$-value for our GPA coefficient is small (0.000553) we can therefore conclude that we reject our null hypothesis that the GPA coefficient is zero, ie. that the GPA term is worth keeping in the model.

A totally equivalent alternative would be to instead create the approximate 95\% confidence interval for the GPA coefficient, by using $\hat{\beta}_j\pm 1.96\text{se}(\hat{\beta}_j)$:

```{r}
#| label: r-medgpa-hypoth-interval
5.454-1.96*1.579
5.454+1.96*1.579
```

The resulting interval is $(2.36,8.55)$, and since it does not include zero, we conclude that the GPA coefficient is significant.

:::
<!-- end of Example -->
<!-- \begin{example}[MLE for the normal linear model] -->
<!-- Consider the model $E(Y_i)= \mu_i = \boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}$ where $Y_i \sim N(\mu_i, \sigma^2)$  for $i =1, \dots, n$ and $\boldsymbol{\beta}$ is a $p$-dimensional vector of parameters. This is a GLM with the identity link. Since $\eta_i =g(\mu_i)=\mu_i$ we have $g'(\mu_i)=1$. The elements of the information matrix therefore are $\mathcal{I}_{jk}=\sum_{i=1}^n \frac{x_{ij}x_{ik}}{\mathrm{Var}(Y_i) (g'(\mu_i))^2} =\sum_{i=1}^n \frac{x_{ij}x_{ik}}{\sigma^2}$ Hence $$
\begin{aligned} -->
<!--               \mathcal{I} =\frac{1}{\sigma^2} \boldsymbol{X}^\intercal \boldsymbol{X} \hfill \text{[label: eqn:informationnormal]} -->
<!--              \end{aligned}
$$
 -->
<!-- Similarly -->
<!--  $\eta_i+z_i = \mu_i+ (y_i-\mu_i) g'(\mu_i)=y_i$ -->
<!--  The estimating equation  $\boldsymbol{X}^\intercal \boldsymbol{W} \boldsymbol{X} \hat{\boldsymbol{\beta}}^{(m)}= \boldsymbol{X}^\intercal \boldsymbol{W} (\boldsymbol{\eta}+\boldsymbol{z}) $ thus simplifies to  $\frac{1}{\sigma^2}\boldsymbol{X}^\intercal \boldsymbol{X} \hat{\boldsymbol{\beta}}=\frac{1}{\sigma^2} \boldsymbol{X}^\intercal \boldsymbol{y} \Rightarrow \hat{\boldsymbol{\beta}}=(\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}. $ -->
<!--  If we write the normal linear model as -->
<!--   $\boldsymbol{y} \sim MVN (\boldsymbol{X}\boldsymbol{\beta}, \sigma^2 \boldsymbol{I})$ where $\boldsymbol{I}$ is the $n\times n$ identity matrix, we have that $E(\hat{\boldsymbol{\beta}}) = E[(\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}]=  (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{X}\boldsymbol{\beta} = \boldsymbol{\beta}, $ -->
<!--  so $\hat{\boldsymbol{\beta}}$ is an unbiased estimator of $\boldsymbol{\beta}$. -->

<!--  To obtain the variance-covariance matrix for $\hat{\boldsymbol{\beta}}$ we use $$(\hat{\boldsymbol{\beta}}-\boldsymbol{\beta}) &= (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}-\boldsymbol{\beta} \\                                                                       &=    (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal (\boldsymbol{y}-\boldsymbol{X}\boldsymbol{\beta}) $$  -->

<!--  Hence, -->
<!--  $$
\begin{aligned} -->
<!--  & E[(\hat{\boldsymbol{\beta}}-\boldsymbol{\beta})  (\hat{\boldsymbol{\beta}}-\boldsymbol{\beta})\intercal] \\ -->
<!--  & = (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal E[(\boldsymbol{y}-\boldsymbol{X}\boldsymbol{\beta})(\boldsymbol{y}-\boldsymbol{X}\boldsymbol{\beta})\intercal]\boldsymbol{X}  (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1}\\ & = (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \mathrm{Var}(\boldsymbol{y})  \boldsymbol{X}  (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1}  \\   &= \sigma^2      (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1}\\ -->
<!--  &= \mathcal{I}^{-1} \hspace{5cm} \text{ by (*{ref{eqn:informationnormal}*)} \end{aligned}
$$
 -->
<!-- In this case, the exact sampling distribution of $\hat{\boldsymbol{\beta}}$ is $N(\boldsymbol{\beta}, \mathcal{I}^{-1})$ or equivalently $(\hat{\boldsymbol{\beta}}-\boldsymbol{\beta})^\intercal \mathcal{I}(\hat{\boldsymbol{\beta}}-\boldsymbol{\beta}) \sim \chi^2(p)$. -->
<!-- \end{example} -->


### Deviance

To assess the adequacy of a model of interest, we compare it with the [saturated]{.term} (or [full]{.term}) model, which has the maximum number of parameters that can be estimated. For data with $n$ observations, $y_1,\dots, y_n$, each with a different parameter in $\boldsymbol{X}^\intercal \boldsymbol{\beta}$, the saturated  model can be specified with $n$ parameters. 

:::{.Key .unnumbered}
Recall that for our GLM we are using @eq-glm-sys to determine our best $\beta$ values, and that we have one equation for each such $i$ in $1,2,\ldots,n$. So our $\beta_1, \beta_2, \ldots, \beta_p$ coefficients will typically be being chosen to try and match the observed means $\mu_i$ of our $y_i$ values simultaneously across $n$ equations. A [saturated]{.term} model includes as many coefficients as equations and thus allows for 'perfect' selection of all the $\mu_i$ values simultaneously. We still have uncertainty as our model for $Y_i$ remains random, and note that having more than $n$ coefficients doesn't help further, it's as if we can already essentially manually select all of our $\mu_i$ values.
:::

If we have replicates (i.e. duplicated $\boldsymbol{x}$ rows), the maximum number of parameters in the saturated model can be less than $n$. 

For the [saturated model]{.term} which uses $m$ parameters (normally $m=n$) we write $\boldsymbol{\beta}_{\max}$ and $\hat{\boldsymbol{\beta}}_{\max}$ for the corresponding parameter vector and its MLE. 

Then let $L(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})$ be the likelihood evaluated at $\hat{\boldsymbol{\beta}}_{\max}$, i.e.\ the likelihood for the [saturated]{.term} model; and let $L(\hat{\boldsymbol{\beta}};\boldsymbol{y})$ be the maximum value of the likelihood for our model of interest (where we used $p$ covariates).

The [likelihood ratio]{.term} 
$$
\lambda=\frac{L(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})}{L(\hat{\boldsymbol{\beta}};\boldsymbol{y})}
$$
provides a measure of how well the model of interest fits compared with the full model.
In practice, we often use the logarithm of the likelihood ratio (since it has more neatly provable asymptotic properties),
$$
\log \lambda = l(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})-l(\hat{\boldsymbol{\beta}};\boldsymbol{y}).
$$

Large values of $\log \lambda$ suggest that the model of interest is a poor description of the data relative to the full model. Note the top of the fraction is always at least as large as the bottom so the minimum of $\lambda$ is $1$. How large a value of $\log \lambda$ is large? To answer this question we can obtain a critical region using the sampling distribution of $\log \lambda$. In fact, we will work with the quantity $2\log \lambda$, which we call the [deviance]{.term}.
 

:::{.Definition}

## Deviance

The deviance, $D$, is defined as 
$$
D=2\log \lambda =2\left[ l(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})-l(\hat{\boldsymbol{\beta}};\boldsymbol{y})\right]
$$
where $l(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})$ is the maximised log-likelihood for the saturated model and $l(\hat{\boldsymbol{\beta}};\boldsymbol{y})$ is the maximised log-likelihood for the model of interest.
:::
<!-- end of Definition -->



<!-- Recall the Taylor series expansion for the log-likelihood (*{ref{eqn:likelihoodexp}*): $(\boldsymbol{\beta}) \approx l(\hat{\boldsymbol{\beta}}) + (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}})^\intercal U(\hat{\boldsymbol{\beta}}) -  \frac{1}{2} (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}})^\intercal \mathcal{I}(\hat{\boldsymbol{\beta}})(\boldsymbol{\beta}-\hat{\boldsymbol{\beta}}) $ -->
<!-- Since $\hat{\boldsymbol{\beta}}$ is the MLE, $U(\hat{\boldsymbol{\beta}})=\boldsymbol{0}$ and hence this becomes  $l(\boldsymbol{\beta})- l(\hat{\boldsymbol{\beta}}) \approx  -  \frac{1}{2} (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}})^\intercal \mathcal{I}(\hat{\boldsymbol{\beta}}) (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}}).$ -->
<!-- Therefore, the statistic $2[ l(\hat{\boldsymbol{\beta}};\boldsymbol{y})-l(\boldsymbol{\beta};\boldsymbol{y})] \approx  (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}})^\intercal \mathcal{I}(\hat{\boldsymbol{\beta}}) (\boldsymbol{\beta}-\hat{\boldsymbol{\beta}})$ will be approximately $\chi^2(p)$. -->
<!-- For the deviance, write $$
\begin{aligned}D&=2[ l(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})-l(\hat{\boldsymbol{\beta}};\boldsymbol{y})]\\ -->
<!--  &=2[ l(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})-l(\boldsymbol{\beta}_{\max};\boldsymbol{y})]\\ -->
<!--  & \hspace{1cm} -2[ l(\hat{\boldsymbol{\beta}};\boldsymbol{y})-l(\boldsymbol{\beta};\boldsymbol{y})]+2 \left[ l(\boldsymbol{\beta}_{\max};\boldsymbol{y})-l(\boldsymbol{\beta};\boldsymbol{y}) \right]    .                    \end{aligned}
$$
 -->
<!--  $2[ l(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})-l(\boldsymbol{\beta}_{\max};\boldsymbol{y})]$ has the $\chi^2(m)$ distribution, where $m$ is the number of parameters in the full (saturated) model. -->
<!--  $2[ l(\hat{\boldsymbol{\beta}};\boldsymbol{y})-l(\boldsymbol{\beta};\boldsymbol{y})]$ is $\chi^2(p)$, where $p$ is the number of parameters in the model of interest. -->
<!--  $v=2 \left[ l(\boldsymbol{\beta}_{\max};\boldsymbol{y})-l(\boldsymbol{\beta};\boldsymbol{y}) \right]$ is a constant which should be close to zero if the model of interest fits the data almost as well as the full model. -->

We won't go into the theory here, but this is a [likelihood-ratio test]{.term}, which you have likely studied elsewhere. Under various conditions (especially large $n$) this quantity $D$ will have an approximately $\chi^2$ distribution, with parameter equal to the difference in degrees of freedom between the two models considered.

This means that for a GLM that fits the data well, the approximate distribution of the deviance, $D$, is $\chi^2_{m-p}$ where $m$ is the number of parameters in the saturated model and $p$ is the number of parameters in the model of interest. We can therefore use this approximate distribution to evaluate whether the assumption here that our model does indeed fit the data well, was reasonable.

We are implicitly assuming that our top level [saturated]{.term} model is at least reasonable, in that our distributional assumptions for the $Y_i$ are correct. Beyond this, the deviance allows us to see whether the restriction placed on our $\eta_i$ values through limiting ourselves to just $p$ covariates still does a good job of fitting the data.

:::{.Example}

## Deviance for a binomial model

Suppose $Y_1, \dots, Y_n$ are independent with $Y_i \sim \mathrm{Bin}(n_i, p_i)$ for $i=1,\dots, n$.  The likelihood function looks like this

$$
L(\boldsymbol{\beta}; \boldsymbol{y}) = \prod_{i=1}^n \binom{n_i}{y_i} p_i^{y_i} (1 - p_i)^{n_i - y_i}.
$$

So the log-likelihood is  

$$
l(\boldsymbol{\beta}; \boldsymbol{y}) = \sum_{i=1}^n \left[y_i \log p_i-y_i \log(1-p_i)+ n_i \log(1-p_i) +\log \binom{n_i}{y_i} \right]
$$ {#eq-binomloglik}


For the full model all the $p_i$'s can be separately determined, so $\boldsymbol{\beta}=(p_1, \dots, p_n)^\intercal$ and with algebra (which we'll skip) it can show that the optimal/maximizing choices are $\hat{p}_i=y_i/n_i$. This gives
  
$$
l(\hat{\boldsymbol{\beta}}_{\max}; \boldsymbol{y}) = \sum \left[y_i \log \left(\frac{y_i}{n_i}\right)-y_i \log\left(\frac{n_i-y_i}{n_i}\right) + n_i \log\left(\frac{n_i-y_i}{n_i}\right) +\log \binom{n_i}{y_i} \right].
$$ {#eq-binomloglikfullmodel}


For any model with $s<n$ parameters, the mle for $p_i$, which we call $\hat{p}_i$ will come from the fitted values $\hat{y}_i$ from the linear model. In the saturated model we were able to ensure $\hat{y}_i=y_i$ for all $i$, but now we likely cannot. So in @eq-binomloglik we replace $p_i$ with $\hat{p}_i = \frac{\hat{y}_i}{n_i}$ to obtain

$$
l(\hat{\boldsymbol{\beta}}; \boldsymbol{y}) = \sum_{i=1}^n \left[y_i \log \left(\frac{\hat{y}_i}{n_i}\right)-y_i \log\left(\frac{n_i-\hat{y}_i}{n_i}\right)+n_i \log\left(\frac{n_i-\hat{y}_i}{n_i}\right) +\log \binom{n_i}{y_i} \right].
$$

Thus, the deviance is 

$$
D = 2[ l(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})-l(\hat{\boldsymbol{\beta}};\boldsymbol{y})]= 2\sum_{i=1}^n \left[ y_i \log\left(\frac{y_i}{\hat{y}_i}\right)+ (n_i - y_i)  \log\left(\frac{n_i-y_i}{n_i-\hat{y}_i}\right) \right].
$$ {#eq-binomdev}


:::
<!-- end of Example -->


:::{.Example}

## Deviance for a normal linear model

Suppose $Y_1, \dots, Y_n$ are independent with $Y_i \sim N(\mu_i, \sigma^2)$ and $E(Y_i)= \mu_i=\boldsymbol{X}^\intercal_i  \boldsymbol{\beta}$ for $i=1,\dots, n$ .
 The likelihood function is
 
 $$
L(\boldsymbol{\beta};\boldsymbol{y}) = \prod_{i=1}^n \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{(y_i-\mu_i)^2}{2\sigma^2} \right).
 $$
 
 So, with a little algebra, the log-likelihood function is
 
$$
l(\boldsymbol{\beta}; \boldsymbol{y}) = -\frac{1}{2\sigma^2}\sum_{i=1}^n(y_i -\mu_i)^2 -\frac{1}{2}n \log(2\pi \sigma^2).
$$ {#eq-normalloglik}


For the saturated model, with $n$ parameters, we can essentially control all the $\mu_1, \dots, \mu_n$. The MLEs are $\hat{\mu}_i=y_i$ and so the maximum value of the log-likelihood becomes
 

$$
l(\hat{\boldsymbol{\beta}}_{\max}; \boldsymbol{y}) =  -\frac{1}{2}n \log(2\pi \sigma^2).
$$ {#eq-normloglikfullmodel}


 For any other model with $p<n$ parameters, the MLE of $\boldsymbol{\beta}$ is $\hat{\boldsymbol{\beta}} = (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}.$
The corresponding maximised log-likelihood function is
 
$$
l(\hat{\boldsymbol{\beta}}; \boldsymbol{y}) = -\frac{1}{2\sigma^2}\sum_{i=1}^n(y_i -\boldsymbol{x}_i^{\intercal} \hat{\boldsymbol{\beta}})^2 -\frac{1}{2}n \log(2\pi \sigma^2).
$$

 The deviance, $D = 2\left[l(\hat{\boldsymbol{\beta}}_{\max};\boldsymbol{y})-l(\hat{\boldsymbol{\beta}};\boldsymbol{y})\right]$ is therefore 
$$
D = \frac{1}{\sigma^2}\sum_{i=1}^n \left(y_i-\boldsymbol{x}_i^{\intercal} \hat{\boldsymbol{\beta}}\right)^2  =\frac{1}{\sigma^2} \sum_{i=1}^n \left(y_i-\hat{\mu}_i\right)^2.
$$ {#eq-normaldev}

              


<!-- In the case when $E(Y_i)=\mu$ (null model), we have $\hat{\mu}_i=\hat{\mu}=\bar{y}$ and -->
<!-- $$
\begin{aligned} -->
<!--   D =\frac{1}{\sigma^2} \sum_{i=1}^n \left(y_i-\bar{y}\right)^2. -->
<!-- \end{aligned}
$$
 -->
<!-- Note the relationship to the sample variance $S^2$: -->
<!-- $S^2 = \frac{1}{n-1}\sum_{i=1}^n \left(y_i-\bar{y}_i\right)^2=\frac{\sigma^2 D}{n-1}.$ -->

<!-- The distribution of the sample variance is given by $(n-1)S^2/\sigma^2 \sim \chi^2(n-1),$ so $D \sim \chi^2(n-1).$ -->

<!-- More generally, when $\mu_i=\boldsymbol{x}_i^{\intercal} \hat{\boldsymbol{\beta}}$, we have -->
<!-- $$
\begin{aligned} -->
<!--   D =\frac{1}{\sigma^2} \sum_{i=1}^n \left(y_i-\boldsymbol{x}_i^{\intercal} \hat{\boldsymbol{\beta}}\right)^2=\frac{1}{\sigma^2} (\boldsymbol{y}-\boldsymbol{X} \hat{\boldsymbol{\beta}})^\intercal (\boldsymbol{y}-\boldsymbol{X} \hat{\boldsymbol{\beta}}). -->
<!-- \end{aligned}
$$
 -->
<!-- Since the MLE is $\hat{\boldsymbol{\beta}}= (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}$, the term $(\boldsymbol{y}-\boldsymbol{X} \hat{\boldsymbol{\beta}})$ can be written as -->
<!-- $$
\begin{aligned}\boldsymbol{y}-\boldsymbol{X} \hat{\boldsymbol{\beta}} &= \boldsymbol{y}-\boldsymbol{X} (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal \boldsymbol{y}\\ -->
<!--  &= [\boldsymbol{I} - \boldsymbol{X} (\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal]\boldsymbol{y}=[\boldsymbol{I}-\boldsymbol{H}]\boldsymbol{y} -->
<!-- \end{aligned}
$$
 where $\boldsymbol{I}$ is the identity matrix and $\boldsymbol{H}= \boldsymbol{X}(\boldsymbol{X}^\intercal \boldsymbol{X})^{-1} \boldsymbol{X}^\intercal$ is the \term{hat} matrix. -->
<!-- Therefore -->
<!-- $$
\begin{aligned} -->
<!--   D &=\frac{1}{\sigma^2} (\boldsymbol{y}-\boldsymbol{X} \hat{\boldsymbol{\beta}})^\intercal (\boldsymbol{y}-\boldsymbol{X} \hat{\boldsymbol{\beta}})\\ -->
<!--   &=\frac{1}{\sigma^2}\left([\boldsymbol{I}-\boldsymbol{H}]\boldsymbol{y}\right)^\intercal [\boldsymbol{I}-\boldsymbol{H}]\boldsymbol{y}\\ -->
<!--   &= \frac{1}{\sigma^2} \boldsymbol{y}^\intercal (\boldsymbol{I}-\boldsymbol{H})\boldsymbol{y} -->
<!-- \end{aligned}
$$
 -->
<!-- since $\boldsymbol{H}$ is idempotent; that is $\boldsymbol{H}^\intercal = \boldsymbol{H}$ and $\boldsymbol{H}\boldsymbol{H}=\boldsymbol{H}$. -->
<!-- $D  = \dfrac{1}{\sigma^2} \boldsymbol{y}^\intercal (\boldsymbol{I}-\boldsymbol{H}) \boldsymbol{y}$ is a quadratic form, with rank equal to the rank of the matrix $\boldsymbol{I}-\boldsymbol{H}$. -->
<!-- The rank of $\boldsymbol{I}$ is $n$ and that of $\boldsymbol{H}$ is $p$, so the rank of $\boldsymbol{I}-\boldsymbol{H}$  is $n-p$. (For more details consult your Predictive Modelling notes.) -->



It turns out that the (exact) distribution of $D$ is $\chi^2(n-p)$.
If the model fits the data well, then $D \sim \chi^2(n-p)$, and the expected value of $D$ will be $n-p$, since the expectation of a chi-squared random variable is equal to its degrees of freedom. 

However, we are not able to use this chi-squared distribution directly, because the expression for the deviance contains the nuisance parameter $\sigma^2$. As you may remember from your linear regression courses, we end up using F tests instead.



<!-- From the expression $D =\frac{1}{\sigma^2} \sum_{i=1}^n \left( y_i-\hat{\mu}_i\right)^2,$ we can obtain an estimate of $\sigma^2$: -->
<!--  $\tilde{\sigma}^2=\frac{ \sum_{i=1}^n \left( y_i-\hat{\mu}_i\right)^2}{n-p}.$ -->

:::
<!-- end of Example -->



:::{.Example}

## Deviance for a Poisson model


Let $Y_1, \dots, Y_n$ be independent random variables with $Y_i \sim Po(\mu_i)$. Then the likelihood function is

$$
L(\boldsymbol{\beta}; \boldsymbol{y}) = \prod_{i=1}^n \frac{\mu_i^{y_i}e^{-\mu_i}}{y_i!},
$$

and so the log-likelihood function is

$$
l(\boldsymbol{\beta}; \boldsymbol{y}) = \sum y_i \log \mu_i - \sum \mu_i -\sum \log (y_i!).
$$ {#eq-poissonloglik}

For the full model $\boldsymbol{\beta}_{\max}= (\mu_1,\dots, \mu_n)^\intercal$, $\hat{\mu}_i=y_i$, and the maximum value of the log-likelihood is

$$
l(\hat{\boldsymbol{\beta}}_{\max}; \boldsymbol{y}) = \sum y_i \log y_i - \sum y_i -\sum \log (y_i!).
$$ {#eq-poissonloglikfullmodel}


Suppose that for the model of interest with $p<n$ parameters the MLE, $\hat{\boldsymbol{\beta}}$, can be used to obtain $\hat{\mu}_i$ and hence fitted values $\hat{y}_i=\hat{\mu}_i$ (because $E(Y_i)=\mu_i$).
 For the model of interest the maximum value of the log-likelihood is
$$
l(\hat{\boldsymbol{\beta}}; \boldsymbol{y}) = \sum y_i \log \hat{y}_i - \sum \hat{y}_i -\sum \log (y_i!).
$$ {#eq-poissonloglikactualmodel}

  So by subtracting @eq-poissonloglikactualmodel from @eq-poissonloglikfullmodel, the deviance is

$$
D= 2 [l(\hat{\boldsymbol{\beta}}_{\max}; \boldsymbol{y}) -l(\hat{\boldsymbol{\beta}}; \boldsymbol{y}) ]= 2 \left[\sum y_i \log\left(\frac{y_i} {\hat{y}_i}\right)-\sum (y_i-\hat{y_i}) \right]
$$ {#eq-devpoisson}


 For most models $\sum y_i=\sum \hat{y_i}$, so the deviance can be written as
$$
D=2 \sum y_i \log\left(\frac{y_i} {\hat{y}_i}\right)= 2 \sum o_i \log\left(\frac{o_i} {e_i}\right),
$$

  where $o_i$ denotes the observed value and $e_i$ the expected value of $y_i$.
 The deviance can be computed from the data and compared with the $\chi^2(n-p)$ distribution.


Consider the data below for which the $Y_i$ are assumed to be independent observations from a Poisson distribution.

<!--
 \begin{center}
\begin{tabular}{crrrrrrrrrr}
\hline
$y_i$ & 2 & 3 & 6 & 7 & 8 & 9 & 10& 12 & 15\\
$x_i$ & -1 & -1 & 0 & 0 &0 & 0 & 1 & 1 & 1 \\
\hline
\end{tabular}
\end{center}
-->

+-------+----+----+----+----+----+----+----+----+----+
| $y_i$ | 2  | 3  | 6  | 7  | 8  | 9  | 10 | 12 | 15 |
+=======+====+====+====+====+====+====+====+====+====+
| $x_i$ | -1 | -1 | 0  | 0  | 0  | 0  | 1  | 1  | 1  |
+-------+----+----+----+----+----+----+----+----+----+


We fit a model of the form $\mu_i=\beta_1 + \beta_2 x_i$.
  The fitted values are $\hat{y}_i=\hat{\beta}_1 + \hat{\beta}_2 x_i$ where $\hat{\beta}_1 =7.45163$ and $\hat{\beta}_2= 4.93530$.
 The deviance is $1.8947$, which is small compared with $n-p=7$, indicating no lack of fit. 
:::
<!-- end of Example -->



:::{.Task}

Verify the value 1.8947 for the deviance given in the above example.
:::
<!-- end of Task -->

::::{.Answer}
We need to get the fitted values $\hat{y}_i$ for $i=1,\dots,9$ and then substitute into the expression for the deviance.

<!-- The latex table should be uncommented in case the pdf doesn't compile with the second table. -->

<!-- \begin{center} -->
<!-- \begin{tabular}{crrrrrrrrrr} -->
<!-- \hline -->
<!-- $y_i$ & 2 & 3 & 6 & 7 & 8 & 9 & 10& 12 & 15\\ -->
<!-- $\hat{y}_i$ & 2.51633 & 2.51633 &7.45163 & 7.45163 &7.45163 & 7.45163& 12.38693 & 12.38693 & 12.38693 \\ -->
<!-- $y_i \log\left(\frac{y_i}{\hat{y}_i} \right)$ & -0.45931 & 0.52743 & -1.30004 & -0.43766 & 0.56807 & 1.69913 & -2.14057 & -0.38082 & 2.87115\\ -->
<!-- \hline -->
<!-- \end{tabular} -->
<!-- \end{center} -->

<!-- full table before rounding
| $y_i$ | 2 | 3 | 6 | 7 | 8 | 9 | 10 | 12 | 15 |
|---|---|---|---|---|---|---|---|---|---|
| $\hat{y}_i$ | 2.51633 | 2.51633 | 7.45163 | 7.45163 | 7.45163 | 7.45163 | 12.38693 | 12.38693 | 12.38693 |
| $y_i \log\left(\frac{y_i}{\hat{y}_i} \right)$ | -0.45931 | 0.52743 | -1.30004 | -0.43766 | 0.56807 | 1.69913 | -2.14057 | -0.38082 | 2.87115 |
-->

Values here are displayed across two tables:

| $y_i$ | 2 | 3 | 6 | 7 | 8 |
|---|---|---|---|---|---|
| $\hat{y}_i$ | 2.51633 | 2.51633 | 7.45163 | 7.45163 | 7.45163 |
| $y_i \log\left(\frac{y_i}{\hat{y}_i} \right)$ | -0.45931 | 0.52743 | -1.30004 | -0.43766 | 0.56807 |

| $y_i$ | 9 | 10 | 12 | 15 |
|---|---|---|---|---|
| $\hat{y}_i$ | 7.45163 | 12.38693 | 12.38693 | 12.38693 |
| $y_i \log\left(\frac{y_i}{\hat{y}_i} \right)$ | 1.69913 | -2.14057 | -0.38082 | 2.87115 |

So $\sum_{i=1}^9 y_i \log\left(\frac{y_i} {\hat{y}_i}\right)=0.94738$ and $D=2\sum_{i=1}^9 y_i \log\left(\frac{y_i} {\hat{y}_i}\right)=1.89476$.

Notice that since we only have three distinct $x_i$ values, we only have three distinct $\hat{y}_i$ values. This is the restriction we are imposing by using a linear model (though technically we have replicates so we may have had no choice here), our model must use the same $\hat{y}_i$ whenever $x_i=-1$, thus preventing us from matching $\hat{y}_i$ exactly with each $y_i$ -- which would be ideal to minimize the likelihood.

In this case, since close $x_i$ values has close $y_i$ values, the likelihood after this restriction is not large compared with a case where we could set each $\hat{y}_i=y_i$, so we didn't see a lack of fit using our linear model.

::::
<!-- end of Answer -->


### Hypothesis testing using the deviance

As we've already seen, we can test hypotheses about the $p$-dimensional parameter vector $\boldsymbol{\beta}$ by using the Wald statistic and the asymptotic distribution of the MLE.

Alternatively we can compare [nested]{.term} models $M_0$ and $M_1$ using the difference of their deviances. Let's consider two such nested models, with $q < p$ covariates:

Model $\mathbf{M_0}$
: $H_0: \boldsymbol{\beta}= \boldsymbol{\beta}_0 = (\beta_1, \dots, \beta_q)^\intercal$.

Model $\mathbf{M_1}$
: $H_1: \boldsymbol{\beta} = \boldsymbol{\beta}_1=(\beta_1, \dots, \beta_p)^\intercal$.
 
We test $H_0$ against $H_1$ by considering 
$$
\begin{aligned}
D_0-D_1& =2\left[l(\hat{\boldsymbol{\beta}}_{\max}; \boldsymbol{y}) -l(\hat{\boldsymbol{\beta}}_0; \boldsymbol{y})\right]-2\left[l(\hat{\boldsymbol{\beta}}_{\max}; \boldsymbol{y}) -l(\hat{\boldsymbol{\beta}}_1; \boldsymbol{y})\right] \\
 &=2\left[ l(\hat{\boldsymbol{\beta}}_1; \boldsymbol{y}) -l(\hat{\boldsymbol{\beta}}_0; \boldsymbol{y})\right]
\end{aligned}
$$

 If both models describe the data well then $D_0 \sim \chi^2(n-q)$, $D_1 \sim \chi^2(n-p)$ and $D_0-D_1 \sim \chi^2(p-q)$.
 If $M_1$ describes the data well but $M_0$ does not, then $D_0-D_1$ will be larger than expected for a value from $\chi^2(p-q)$.
 So, reject $H_0$ if $D_0-D_1> \chi^2(1-\alpha; p-q)$ that is, if the difference in deviances exceeds the upper $100\times \alpha \%$ point of the $\chi^2(p-q)$ distribution.



:::{.Example}

## Hypothesis test for the GPA coefficient in the model for medical school admissions, this time using deviances


Suppose that we want to test $H_0: \beta_1=0$ in the medical school admissions example. We can perform this test using the deviances given in the output.

```{r echo=FALSE}
#| label: r-medgpa-hypoth-code-call
summary(med.glm)
```

Here $D_0$ is the null deviance, that is the deviance in the model that includes only the intercept (and no other predictors). $D_1$ is the residual deviance, that is the deviance of the model of interest (the model with GPA included as a predictor). Under $H_0$, $D_0-D_1$ should be approximately distributed as $\chi^2_1$. The 95th percentile of the $\chi^2(1)$ distribution, sometimes written $\chi^2(0.95; 1)$, is $3.84$, and since $D_0-D_1=75.791-56.839=18.952>3.84$ we can reject the null hypothesis. Again we conclude that GPA is a significant term in the model.



:::
<!-- end of Example -->

<!-- \begin{example}[Normal linear model] -->
<!-- Consider the model $E(Y_i)=\mu_i= \boldsymbol{x}_i^{\intercal} \boldsymbol{\beta}, \hspace{1cm} Y_i \sim N(\mu_i, \sigma^2)$ where the $Y_i, i=1,\dots, n$ are independent. -->
<!-- Suppose that model $M_1$ has $p$ parameters and model $M_0$ has $q$ parameters and that the fitted values from each model are denoted by $\hat{\mu}_i(1)$ and  $\hat{\mu}_i(0)$ respectively. -->
<!--  Then -->
<!--  $ D_0 =\frac{1}{\sigma^2} \sum_{i=1}^n \left[ y_i-\hat{\mu}_i(0)\right]^2$ and -->
<!--  $ D_1 =\frac{1}{\sigma^2} \sum_{i=1}^n \left[ y_i-\hat{\mu}_i(1)\right]^2.$ -->
<!--  Assuming that model $M_1$ fits the data well, we have $D_1 \sim \chi^2(n-p)$. If also $M_0$ fits the data well, then $D_0 \sim \chi^2(n-q)$ and $D_0-D_1 \sim \chi^2(p-q)$. -->
<!--  Since the deviance for the normal model involves the unknown parameter $\sigma^2$, use the ratio -->
<!--   $$
\begin{aligned} -->
<!--    F&=\frac{(D_0-D_1)/(p-q)}{D_1/(n-p)}\\ &= \frac{\left\lbrace \sum_{i=1}^n \left[ y_i-\hat{\mu}_i(0)\right]^2-\sum_{i=1}^n \left[ y_i-\hat{\mu}_i(1)\right]^2 \right\rbrace/(p-q)}{\sum_{i=1}^n \left[ y_i-\hat{\mu}_i(1)\right]^2/(n-p)} -->
<!--   \end{aligned}
$$
 -->
<!-- which should be distributed as $F(p-q,n-p)$ if $H_0$ holds. -->
<!-- \end{example} -->

<!-- \begin{example}[Inference for birthweight v. gestational age data] -->
<!-- Recall Example 1.3 on birthweight as a function of gestational age. -->
<!-- \begin{figure}[hbtp] -->
<!-- \centering -->
<!-- \includegraphics[totalheight=0.8\textwidth,angle=0]{birthweight} -->
<!-- \caption{Birthweight against gestational age for boys (solid circles) and girls (open circles)} \hfill \text{[label: fig:birthweight]} -->
<!-- \end{figure} -->

<!-- $M_0$ is the model under $H_0$, specified as -->
<!-- $E(Y_{jk}) = \mu_{jk}=\alpha_j + \beta x_{jk} ;  \hspace{1cm} Y_{jk} \sim N(\mu_{jk},\sigma^2)$ -->
<!-- $M_1$ is the model under $H_1$, specified as: -->
<!-- $E(Y_{jk}) = \mu_{jk}=\alpha_j + \beta_j x_{jk}; \hspace{1cm}  Y_{jk} \sim N(\mu_{jk},\sigma^2)$ -->

<!-- where the $Y_{jk}$ are independent for $j=1,2$ and $k=1,\dots, 12$. -->
<!-- \begin{center} -->
<!-- \begin{tabular}{cccc} -->
<!-- \hline -->
<!--  Model  & Min. Sum of Squares & No. parameters & DF\\ -->
<!--  \hline -->
<!--  $M_0$ & $\hat{S}_0=658770.8$ & 3& 24-3=21\\ -->
<!--   $M_1$& $\hat{S}_1=652424.5$ &4 & 24-4=20\\ -->
<!--  \hline -->
<!-- \end{tabular} -->
<!-- \end{center} -->

<!-- The deviances are related to the sums of squares: -->
<!--   $\hat{S}_0= \sigma^2 D_0 \Rightarrow D_0=\hat{S}_0/\sigma^2 \text{ and similarly } D_1=\hat{S}_1/\sigma^2$ -->
<!-- Therefore $ F=\frac{(658770.8-652424.5)/1}{652424.5/20}=0.19 $ which is not significant when compared to the $F(1,20)$ distribution. -->
<!-- \end{example} -->

<!-- # Summary -->
<!-- Key points from this chapter: -->

<!-- *  Approximate (asymptotic/large sample) distribution for the score -->
<!-- *  Approximate (asymptotic/large sample) distribution for the MLE -->
<!-- *  Terminology: full (saturated, maximal) model; null (minimal) model; model of interest (something in-between) -->
<!-- *  maximised log-likelihood for full model, model of interest, null model -->
<!-- *  Definition of deviance, deviance for model of interest, deviance for null model -->
<!-- *  Approximate (asymptotic/large sample) distribution for the deviance of a model of interest under the hypothesis that the model is a good fit -->
<!-- *  What is different about the normal linear model and why don't we use the deviance directly for goodness of fit/hypothesis tests? -->
<!-- *  Hypothesis tests about model parameters: Wald tests and tests based on deviances: what are the relevant null hypotheses, test statistics and distributions involved? Can you perform both types of tests for a given example? -->

